---
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

# Upper Extremity and Thoracic Outlet

**21 Jan 2021**: *Kush Sharma, MD and Ashraf Mansour, MD*

```{r echo=FALSE}

library(embedr)

embed_audio("https://traffic.libsyn.com/audiblebleeding/VSITE_Upper_Extremity_post_auphonic.mp3")

```

## Anatomy/ Exposure of Vessels

**What are the zones of the upper extremity?**
[@illig57UpperExtremity2019; @illig57UpperExtremity2019]

Division of the upper extremity into three zones:

1.  Intrathoracic zone including aortic arch, innominate artery,
    subclavian artery bilaterally, innominate veins, and SVC

2.  Thoracic outlet (base of neck to the axilla including the
    subclavian, proximal vertebral, proximal axillary arteries/veins)

3.  Axilla to fingers (the arm)

**What are some common exposures for major upper extremity arteries?**

Right Subclavian Artery: Medial sternotomy (proximal) or right
supraclavicular area (mid/distal)

Left Subclavian Artery: Anterolateral thoracotomy in emergent setting
for proximal left subclavian artery control. When third space
sternotomy, supraclavicular incision with thoracotomy "trap door"
exposure

Supraclavicular incision: After division of the platysma and clavicular
head of the SCM, fat pat of varying thickness contains the omohyoid
muscle. This should be divided and placed superiorly/laterally. At this
point, the anterior scalene muscle is exposed medially with phrenic
nerve running in lateral to medial direction. Division of anterior
scalene for carotid/subclavian bypass should be performed as close to
the first rib as possible. After this is performed, the subclavian
artery is exposed.

Axillary Artery: Infraclavicular exposure below middle 1/3rd of
clavicle. Pec major split and pec minor freed at lateral wound. Axillary
vein followed by deep and superior to get to artery

Anatomically bound by the first rib proximally and the lateral edge of
the teres major muscle distally. For exposure of the first part of the
axillary artery, the ipsilateral arm is abducted approximately 90
degrees and horizontal skin incision 2 cm below the middle third of the
clavicle. Underlying pec major is split by bluntly separating the fibers
and followed by exposing the tough clavipectoral fascia. At the lateral
wound, the pec minor can be freed and laterally retracted. The axillary
vein is first structure encountered in the sheath and the artery lies
just superior and deep to the vein. Make sure to avoid nerves of
brachial plexus that lie deep to first part of axillary artery and are
at risk for injury during blind placement of occluding arterial clamps.
[@garygwindAnatomicExposuresVascular2013]

**What steps are involved for brachial artery exposure?**

Brachial artery: incision between biceps/triceps on medial arm (avoid
basilic vein damage in subcutaneous and deep to the fasia at medial
biceps. Median nerve seen and retracted. Two brachial vein are paired
adjacent to artery.

Superficial location makes it vulnerable to injury and accounts for most
vascular injuries of upper extremities. Brachial artery exposure
involves a 5-8 cm longitudinal incision in the groove between the
biceps/triceps muscles on the medial aspect of the arm. In the lower
half of the arm, take care to avoid basilic vein damage in the
subcutaneous tissue. Neurovascular bundle exposed by incising the deep
fascia at the medial border of the biceps muscle, which is retracted
anteriorly. After retracting basilic vein into posterior wound ,brachial
sheath is opened and median nerve is most superficial structure and
retracted. The artery lies deep to the nerve and surrounded by two
brachial veins. Posteriorly, is the presence of the ulnar nerve.

Brachial Artery bifurcates at the radial tuberosity into radial/ulnar
branches. After the bifurcation and immediately after its origin, the
ulnar artery gives off a short common interosseous branch, which
bifurcates at the hiatus in the proximal interosseous membrane. Exposure
of brachial artery in the antecubital fossa requires a transverse skin
incision 1 cm distal to the midpoint of the antecubital crease. After
deepening, avoid injury to subcutaneous veins and mobilize the basilic
vein medially. Medial antebrachial cutaneous nerve should be protected.
Divide the bicipital aponeurosis and after division, exposure of the
brachial artery is present, which is flanked by two deep veins and
crossing branches. Isolation of brachial artery requires ligation and
division of these crossing vein branches.

Radial artery at the wrist with 2-3 cm longitudinal incision generally
between radial artery and cephalic vein. Radial artery was exposed by
incising the antebrachial fascia just medial to the radius. Two veins
accompany the artery and should be dissected away during arterial
isolation. The superficial radial nerve and its medial/lateral branches
course between the cephalic vein and radial artery in the area.

Exposure of the ulnar artery is by coursing beneath the superficial
flexor muscles in the proximal forearm, emerging near the ulnar border
at the point midway between the elbow and the wrist. In the distal
forearm, the ulnar artery course just beneath the antebrachial fascia
and is easily exposed through a longitudinal incision placed radial to
the flexor carbi ulnaris. The palmar branch of the ulnar nerve courses
the superficial to the antebrachial fascia and should be preserved
during arterial exposure

**What common aberrant upper extremity/arch anatomy is important to be
aware of?**

-   Bovine arch with left common carotid/left subclavian have common
    origin

-   Vertebral artery directly off the aortic arch

-   Aberrant right subclavian where innominate becomes right CCA and
    right subclavian distal to last branch on left side passing behind
    esophagus to supply the right arm

## Epidemiology, etiology, and diagnostic evaluation

**How does evaluation of upper extremity ischemia differentiate from
lower extremity ischemia?** [@shuja117UpperExtremity]

-   Upper extremity ischemia \<5% of patients with limb ischemia and in
    contrast to lower extremity, atherosclerosis is not a major
    contributor to upper extremity ischemia

-   Vast majority of cases caused by autoimmune/connective tissue
    disorders

**How can upper extremity disease be classified?**

Anatomic Location:

-   Large vs. Small Vessel

Disease Process:

-   Vasospastic or occlusive. Vasospastic disease is more responsive to
    pharmacologic management while occlusive requiring
    endovascular/surgical management.

**How should patients be evaluated who have concern for upper extremity
disease?**

Diagnostic Evaluation

1.  Detailed H+P evaluation (pulse palpation, auscultation at
    supraclavicular/infraclavicular fossa may reaveal a bruit concerning
    for subclavian artery stenosis, upper extremity neurovascular/skin
    exam)

2.  Brachial/forearm blood pressures and if suspected claudication,
    measured at rest and 2-5 minutes after exercise. Look for a gradient
    of >20 mmHg is considered significant

3.  Some or all of 6 P's of acute limb ischemia with symptoms occurring
    within 14 days are deemed acute

4.  Doppler insonation of radial, ulnar, palmar, and digital arteries

5.  Vascular Lab Evaluation

    1.  Segmental Pressure Measurements

    2.  Duplex Ultrasound (look for large vessel occlusive disease)

6.  Other Imaging

    1.  CTA/MRA

7.  Clinical Lab tests

    1.  Inflammatory disorders-CBC, ESR, ANA, RF

    2.  Hypercoagulable screening

## Operations/Procedures

**What are some indications for carotid-subclavian bypass?**

1.  Atherosclerosis 

2.  Staged revascularization prior to TEVAR for aneurysmal disease
    requiring coverage of the LSA

**How does the exposure differentiate in transposition vs bypass?**

Exposure (Transposition vs Bypass)

-   Arterial transposition via a short, transverse cervical incision
    above the clavicle between two heads of SCM (bypass is lateral to
    entire SCM)

-   Sub-platysmal flaps created and avoid EJ vein damage

-   Omohypoid divided between heads of SCM and IJ mobilized laterally
    (bypass IJ is mobilized medially to expose CCA and care must be
    taken to avoid phrenic nerve in more lateral approach)

-   CCA is reflected medially with vagus nerve 

-   On the left side, the thoracic duct is identifiable and divided
    followed by dividing the vertebral vein

-   Subclavian artery and proximal branches identified (anterior scalene
    is in lateral dissection)

**What are some common complications after carotid subclavian bypass in
order of highest to lowest incidence?**

Complications [@voigtOutcomesCarotidsubclavianBypass2019]

1.  Phrenic nerve palsy  

2.  Recurrent laryngeal palsy  

3.  Lymphatic leak  

4.  Neck hematoma  

**When carotid-subclavian bypass compared to transposition?**  

1.  Vertebral artery takes origin from the subclavian artery in a very
    proximal position, then Bypass

2.  For coronary-subclavian steal with patent internal mammary artery to
    coronary artery bypass graft, then Bypass (a carotid-subclavian
    transposition requires a more proximal clamp with occlusion of
    inline antegrade flow to the coronary bypass during the procedure)

## Vaso-occlusive disease

**What are causes and symptoms associated with subclavian/axillary
occlusive disease?** [@jacklcronenwettVascularDecisionMaking2020]

-   Etiology: Atherosclerosis is the most common cause of
    subclavian/axillary occlusive disease. Left SCA > Right involvement.
    Less common causes include Takayasu disease, giant cell arteritis,
    or arterial TOS

-   Symptoms: Upper extremity arm/hand ischemia or neurologic symptoms
    due to subclavian-vertebral steal. Because significant collaterals,
    minimal pain on exertion even with subclavian occlusion

**What are causes and symptoms associated with brachial/forearm
occlusive disease?**

-   Etiology: MCC of brachial artery occlusion is cardiac origin
    embolus. Atherosclerosis RARELY affects the brachial artery. Distal
    axillary/proximal brachial stenosis can be from repetitive trauma
    from crutch use. 

-   Forearm occlusive disease can be seen in advanced ESRD/DM where
    calcific atherosclerosis of radial/ulnar arteries is present. Less
    common causes include Beurger disease or Raynaud Phenomenon 

**How/when is upper extremity occlusive disease treated?**

-   SCA Occlusive Disease

    -   Endovascular: treated with stenting if they are short segment
        and not close to the vertebral artery origin. Balloon expandable
        stents favorable for ostial lesions and access can be via the
        femoral or ipsilateral brachial artery 

    -   Surgery:

        -   Bypass from aortic arch through median sternotomy 

        -   Ipsilateral CCA to subclavian artery (bypass or
            transposition) 

        -   Contralateral CCA (anterior or retropharyngeal)

-   Brachial/forearm Occlusive disease 

    -   Endovascular: PTA evidence is anecdotal with stents for lesions
        unresponsive to PTA or dissection following angioplasty 

    -   Surgery: 

        -   GSV vein bypass remains standard for revascularization with
            bypasses to superficial or deep palmar arch have good
            patency rates. Tunneling is subcutaneous if to distal ulnar
            or superficial palmar arch whereas anatomical to distal
            radial artery over the anatomic snuffbox 

## Vasospastic Disorders

**What is Raynaud's and what causes it?** [@shuja117UpperExtremity;
@landry141RaynaudPhenomenon2019]

-   Exaggeration of normal physiologic response with episodic pallor or
    cyanosis of the fingers caused by small digital artery
    vasoconstriction occurring in response to cold or emotional stress.
    There is an abnormality with sympathetic nervous system, resulting
    in a multifactorial problem involving a combination of vascular,
    neural, and humoral factors.

**What are the subtypes of Raynaud's phenomenon and what is the
underlying pathology?**

-   Primary: Raynaud's disease-idiopathic form that is a benign process
    not associated with structural vascular change. Triggers include
    (cold, emotional stress, caffeine) resulting in digital smooth
    muscle contraction and temporary digital hypoperfusion.

-   Secondary: Fixed vascular obstruction to blood flow decreasing
    threshold for cold induced vasospasm or progress to tissue loss.
    Diseases associated include mixed connective tissue disease, SLE,
    and rheumatoid arthritis, and scleroderma (accounts for 80-90% of
    cases). In setting of lower digital blood pressure, symptomatic
    digital ischemia or tissue loss under low stress conditions. With
    cold/emotional stress, vasoconstrictive response of digital artery
    smooth muscle further causes arterial closure and resultant symptoms

**What are diagnostic criteria for Raynaud's?**

-   Clinical (Progression of ischemia with white -> blue -> red finger
    discoloration. Episodes can be self-limited and may last from less
    than a minute, but generally not longer than 10-20 minutes 

-   Qualitative testing for severity of cold sensitivity in Raynaud's
    syndrome can be useful. Most basic test is cold sensitivity and
    recovery after ice water immersion. >10 minutes return to baseline
    pressure concerning for Raynaud's

-   Segmental pressures with finger systolic blood pressure can
    differentiate purely vasospastic vs occlusive disease. Difference of
    more than 15 mm Hg between fingers or absolute finger pressure \<70
    mm Hg may indicate occlusive disease 

-   Serologic evaluation (ANA/RF)

**What are appropriate treatments for Raynaud's phenomenon?**

1.  Medical-cold/tobacco avoidance. Calcium channel blocker (nifedipine)
    has been the most effective and losartan has also been beneficial.
    Fluoxetine (SSRI). Other drugs include alpha blocker, sildenafil,
    reserpine, cilostazol, captopril. NOT GOOD OUTCOMES IN PATIENTS WITH
    ARTERIAL OBSTRUCTION 

2.  Surgical-thoracic sympathectomy (used for treatment of digital
    artery vasospasm/digital ischemic ulceration). For vasospasm,
    thoracic sympathetcomy is initially successful, but symptoms return
    generally within 3-6 months. 

3.  Immunosuppression/immunomodulation for connective tissue disorders
    associated with secondary Raynaud phenomenon

### **Ergotism**

**What is Ergotism?** [@jamescstanleyCurrentTherapyVascular2014]

-   Etiology: Ergot is a parasitic fungal disease that has a particular
    prevalence for infecting rye plants and ergot alkaloids have been
    linked to epidemic poisonings that manifested as ergotism from
    consumption of rye 

-   Modern day is rare

**What causes Ergotism and how do patients present?**

-   Ergotamine is chemically like endogenous catecholamines/indolamines
    and when applied clinically, it behaves as an agonist to
    alpha-adrenergic, sertoninergic, and dopaminergic receptors. Despite
    limited bioavailability, vasocontrictive effects have been reported
    to last for 24 hours or longer 

-   Gangrenous-mild limb pain followed by burning pain/shooting and 

-   Convulsive-heaviness in limbs and head associated with diarrhea.
    Could result in tonic-clonic spasms 

**How can you diagnose Ergotism and what is the process for treating
this disease?**

Upper extremity ischemia (i.e. digital ulceration) in the setting of
ergot alkaloid use (typically for migraines) 

Treatment: 

-   Volume expansion and IV heparin as anticoagulation 

-   IV infusion of nitroprusside, nitroglycerin, ilioprost or
    combination

-   Infusion of Ca 2+ channel blockers 

-   Surgical: for thrombosis, consider thrombolysis 

### **Buerger's Disease**

**How is Buerger's disease categorized?**
[@jacklcronenwettVascularDecisionMaking2020]

-   Non-atherosclerotic, segmental, inflammatory disease of small/medium
    sized arteries in distal extremities of tobacco users distinct from
    either atherosclerosis of immune arteritis

**What clinical criteria can help diagnose Buerger's?**

-   Smoking history, onset before 50 years, infrapopliteal arterial
    occlusions, upper limb involvement, absence of atherosclerotic risk
    factors besides smoking

**What is important about diagnosing Buerger's**

-   Typically a diagnosis of exclusion

-   Must rule out proximal embolic source, trauma, local lesions (eg pop
    entrapment or cystic adventitial disease), autoimmune disease,
    hypercoagulable status, atherosclerosis

**What physical exam and non-invasive/invasive imaging findings of
Buerger's?**

-   Distal, but not proximal arterial disease (palpable
    brachial/popliteal but absent/reduced at ankle or wrist)

-   DBI\<0.6 and flat/reduced digital waveforms

-   CTA/MRA/DSA-characteristic corkscrew collateral

**What is the mainstay treatment in Buerger's disease?**

1.  Smoking cessation! Only treatment to improve symptoms and reduce
    amputation risk if achieved before onset of gangrene or tissue loss.
    Important to remember following treatments will likely fail without
    smoking cessation. 

2.  If smoking cessation does not improve, medical management with
    antiplatelet agents, immunomodulators, vasodilators, anticoagulants 

3.  Endovascular-distal small vessel intervention

4.  Surgical-upper extremity autogenous vein bypass-limited success due
    to poor outflow 

5.  Sometimes can consider upper extremity sympathectomy, but unproven
    benefit 

6.  Amputation-reported in 30-40% who are followed longer than 5 years 

### **Large Artery Vasculitis**

**What are common characteristics for patients who are suspected to have
a large vessel vasculitis?** [@shanmugam137VasculitisOther2019]

-   Affect aorta and major branches 

-   Present with non-specific heterogenous symptoms making the diagnosis
    challenging. Most commonly, they present with systemic or
    constitutional symptoms (fatigue, fever, weight loss, arthralgias)

-   Frequently, diagnosis made with presence of constitutional symptoms,
    elevated inflammatory markers, and dedicated imaging (MRA, CTA, DUS,
    or PET)

**How can you differentiate takayasu arteritis vs giant cell
arteritis?**

1.  Takayasu arteritis 

    1.  Aorta and primary 

    2.  Young patients \<20 years and female in 80-90% of cases, Asian
        populations

    3.  Criteria (ACR)

        1.  Onset \<40 years

        2.  Claudication of an extremity 

        3.  Decreased brachial pulse 

        4.  \>10 mmHg SBP between arms

        5.  Bruit over subclavian arteries or aorta

        6.  Arteriographic evidence of narrowing/occlusion in
            aorta/primary branches/or large upper/lower extremity
            arteries

2.  Giant cell arteritis 

    1.  Aorta and main branches, but pre-dilection for carotid artery
        branches

    2.  Diagnosis: 

        1.  Age at disease onset > 50 years 

        2.  New headache

        3.  Temporal artery abnormality 

        4.  Elevated ESR (>50) 

        5.  Abnormal artery biopsy (gold standard test)

    3.  Other symptoms include jaw pain with mastication or visual
        changes

    4.  Associated with Polymyalgia rheumatic, characterized by morning
        stiffness in shoulders/hips occurring in 40-50% of patients 

    5.  Arteriography/MRA/CTA/PET may be used to assess large vessel
        involvement

**How should patients be monitored with active large artery
vasculitis?**

-   Lab data tracked at least monthly for 6 months with close follow-up
    to ensure appropriate response to medical treatment and enable
    physicians to assess for adverse effects of medical treatment

-   Repeat tests after remission reached and imaging choice to evaluate
    large vessels (DUS/CTA/MRA)

**What is the medical treatment for GCA and when do you consider
surgical treatment?**

-   Medical-steroid therapy. In as many as 50% of patients who have a
    large vessel vasculitis refractory to glucocorticoid therapy alone,
    patients will trial immunomodulators or cytotoxic truxs (ie
    methotrexate, azathioprine, mycophenolate, tocilizumab, or
    leflunomide) 

-   Intervention-once remission, treatment of symptomatic arterial
    lesions should be considered and as many as 50-70% with large vessel
    vasculitis will require intervention. 

    -   Endovascular-angioplasty/stent/stent graft for large vessel
        vasculitis have all been described, however higher restenosis in
        endovascular compared to open treatment

    -   Open Surgery (gold standard)-lesions are long, fibrotic and
        therefore less amenable to endovascular treatment. Bypass grafts
        from aorta-CCA are the most common (CEA should be avoid due to
        pathology involved)

        -   Upper extremity bypass with autogenous vein to the brachial
            artery

        -   Aortic aneurysms should be managed with open surgery 

## Aneurysmal Disease

**How are subclavian aneurysms caused and how can they present?**
[@baig84UpperExtremity2019]

Etiology/Pathology:

-   Degenerative (atherosclerotic or due to aberrant right subclavian
    with degenerative changes in proximal subclavian known as "Kommerell
    diverticulum")

-   Traumatic (blunt, penetrating, iatrogenic with attempted catheter
    placement) 

-   Thoracic outlet obstruction

Presentation

-   Exam-pulsatile supraclavicular mass or bruit, absent/diminished
    pulses, signs of microembolization ("blue finger")
-   Most discovered incidentally, however referred chest, neck, shoulder
    pain, upper extremity ischemia due to thromboembolic phenomenon,
    brachial plexus compression, hoarseness from right recurrent
    laryngeal nerve compression
-   Dysphagia from esophageal compression in aberrant right subclavian
    artery

**What are diagnostic studies and treatment modalities for subclavian
aneurysms?**

-   CXR-mediastinal mass may suggest neoplasm

-   MRA/CTA important to delineate extent of aneurysm and proximity to
    ipsilateral vertebral artery 

Treatment: 

-   Open Repair-resection/endoaneurysmorrhaphy with end to end (small
    aneurysms) or interposition prosthetic graft 

    -   Proximal-median sternotomy with supraclavicular fossa extension
        for adequate proximal control for right side, however
        supraclavicular with left anterolateral thoracotomy for left
        subclavian aneurysm 

    -   Mid-Distal-supraclavicular/infraclavicular generally adequate
        for control where again resection of the clavicle may be needed 

-   Endovascular Repair-transbrachial/transfemoral approach with covered
    stent 

    -   Must consider vertebral artery origin. Can cover vertebral
        artery if contralateral vertebral artery is patent and of
        adequate size, however posterior circulation stroke may occur
        when the contralateral vertebral artery is highly stenotic,
        hypoplastic or occluded.

-   Hybrid Repair-embolization/coils of proximal subclavian artery
    combined with subclavian transposition or carotid-subclavian bypass

-   For aberrant subclavian artery aneurysm, resection or exclusion of
    the aneurysmal artery with vascular reconstruction of the subclavian
    artery is recommended. Especially in the setting of dysphagia
    lusoria, subclavian artery reconstructed by interposition graft
    where proximal anastomosis is on ascending aorta. Alternatively,
    left posterolateral thoracotomy for proximal aneurysm resection and
    right supraclavicular incision for reconstruction of subclavian
    artery by end to side to the right CCA has been reported.

**How are axillary aneurysms caused and how can they present?**

Etiology/Pathology: 

-   Blunt/penetrating trauma

-   Congenital (infrequently reported)

-   Post-traumatic axillary aneurysms (repeated abduction/external
    rotation downward toward humeral head in baseball pitchers) 

Presentation: 

-   Exam-pulsatile supraclavicular mass or bruit, absent/diminished
    pulses, signs of microembolization ("blue finger")

**What are diagnostic studies and treatment modalities for axillary
aneurysms?**

Diagnosis: 

-   Ultrasound

-   CTA/MRA of upper extremity

Treatment: 

-   Open Repair-resection with interposition vein grafting or prosthetic
    if inadequate vein is present. 

-   Endovascular repair-covered stent graft can be placed with
    occasional embolization with micro coils to isolate sac and prevent
    retrograde endoleaks

**How are brachial aneurysms caused and how can they present?**

Etiology/Pathology: 

-   False aneurysms secondary to repetitive trauma

-   Iatrogenic complications 

-   IV drug abuse (infected pseudoaneurysms in antecubital fossa) 

-   Connective tissue disorders (ex. type IV Ehlers danlos)

Presentation: 

-   Exam: pulsatile mass 

-   Local pain or symptoms of median nerve compressions

-   Hand/digital ischemia from thrombosis/distal embolization

**What are diagnostic studies and treatment modalities for brachial
aneurysms?**

Diagnosis: 

-   Duplex Ultrasound

-   CTA/MRA of upper extremity may be needed to delineate extent of
    aneurysm 

Treatment: 

-   Open Repair (preferred)-resection with patch or interposition vein
    grafting  

-   Endovascular repair-rare and generally in a traumatic setting

-   Iatrogenic injuries-due to access and nonoperative treatment for
    small/asymptomatic pseudoaneurysms that are likely to thrombose
    spontaneously. Direct suture repair with evacuation of hematoma is
    possible. Thrombin injection is less favorable due to location and
    short neck. 

## Occupational Vascular Disease

**There are some occupational vascular disorders than contribute to
vascular disease in the upper extremity. Hand arm vibration syndrome and
hypothenar hammer are of particular importance. Can you talk to us about
the key information from these syndromes?**
[@eskandari185ConditionsArising2020]

### Hand-Arm Vibration Syndrome 

Etiology:

-   Vibrating handheld machines (eg pneumatic hammers and drills,
    grinders, and chain saws) 

-   Linear relationship between exposure over years and onset of this
    syndrome 

-   Exact mechanism unknown, but thought that endothelial damage with
    sympathetic hyperactivity -> finger blanching attack 

Presentation:

-   Various stages seen where early results in slight tingling/numbness
    and lateral, the tips of one or more fingers experience attacks of
    blanching that is usually precipitated by cold 

-   Blanching typically lasts 1 hour and terminates with reactive
    hyperemia, but prolonged exposure can cause bluish black cyanosis of
    fingers 

Diagnosis 

-   Detailed history with use of vibrating tools/symptoms of Raynaud
    phenomenon 

-   Objectively: cold induced ischemia with recording time until digital
    temperature recovers

-   Digital occlusion with noninvasive digit pressures or duplex
    scanning

Treatment

-   Avoidance of vibratory tools

-   Nifedipine (Ca2+ channel blocker) in advanced cases 

-   IV prostanoid (ie prostacyclin) for digital gangrene 

-   Surgery-cervical sympathetectomy or digital sympathectomy rarely
    needed\

### Hypothenar hammer syndrome

Etiology: 

-   Repetitive use of palm of hand in activities that involve pushing,
    pounding, twisting

-   Name comes from reports of mechanics, factory workers, carpenters or
    laborers who habitually use there hands as a hammer are ad risk for
    disease 

-   Repetitive trauma leads to thrombotic occlusion, aneurysm formation
    or both

Presentation: 

-   Asymmetrical distribution involving dominant upper extremity where
    cyanosis and pallor can occur and digits affected are ulnar
    distribution in nature 

-   Cool/mottled digits or severe cases with ischemic ulcers

Diagnosis: 

-   Duplex ultrasound 

-   CTA or MRA 

-   Arteriography (gold standard) with corkscrew pattern typically in
    affected vessels

Treatment 

-   Conservative-smoking cessation/hand protection/cold avoidance 

-   Medical-calcium channel blockers/antiplatelet

-   Surgical (severe digital ischemia/aneurysm)-ligation if adequate
    collateral or interposition vein graft

### Environmental Exposures

**Exposure to what environmental agents can result in upper extremity
ischemia?**

Acrosteolysis

-   Exposure to polyvinyl chloride can result in ischemic hand symptoms
    similar to those of Raynaud syndrome 

-   Angiography-damage to digital arteries with multiple
    stenosis/occlusions or hyper vascularity adjacent to areas of bone
    resorption 

-   Treatment-supportive

Electrical burns

-   \<1000 V cause injuries limited to immediate skin/soft tissue,
    however >1000 V cause damage from entry to exit point 

-   Results in arterial necrosis with thrombus or bleeding and gangrene
    of digits develop 

-   Initially can be occlusion/thrombosis or spasm, however later damage
    can cause aneurysmal degeneration 

-   Treatment-dependent on soft tissue/bone injuries as well. Can have
    reconstruction with free flap due to local vascular damage or
    occlusion of major artery requiring bypass grafting

Extreme thermal injuries

-   Workers at risk with chronic exposure to cold (slaughterhouse,
    canning factory, and fisheries)

-   Raynaud syndrome symptoms due to vasomotor disturbances in the hands
    when exposed to extreme chronic thermal trauma

-   Treatment-Supportive 

### Sports Medicine

**How can athletes specifically be affected by upper extremity
ischemia?**

Overview

-   Athletes who engage in strenuous or exaggerated hand/shoulder
    activity may be susceptible to upper extremity ischemia from
    arterial injury manifested by Raynaud syndrome, symptoms of sudden
    arterial occlusion or digital embolization

## Vascular Trauma-Upper Extremity

**This is discussed in detail here:** \@ref(vascular-trauma), **so we
will go over some important specifics for upper extremity vascular
injury.** [@kauvar184VascularTrauma2020]

**What is the mechanism and management of upper extremity axillary
artery trauma?**

Mechanism and Pattern

-   Predominantly in penetrating trauma with equal incidence in
    proximal/middle/distal divisions and brachial plexus injury
    in >1/3rd of arterial injury

Diagnostic Considerations

-   Physical exam with deficiencies in upper extremity pulses/ischemic
    changes, but may not be present given collateral flow from axillary
    artery to upper extremity 

-   High index of suspicion with location of injury proximity to course
    of axillary artery

-   Upper extremity Doppler or CTA if patient is stable for diagnosis

Surgical Considerations 

-   Primary repair or treated with interposition graft 

-   If hemodynamically stable, can consider covered stent based on
    location to thoracic outlet via femoral/brachial approach

**What is the mechanism and management of upper extremity brachial
artery trauma?**

Mechanism and Pattern

-   Frequently associated with humerus fractures/elbow dislocation

-   Penetrating trauma

Diagnostic Considerations

-   Pulse deficit in majority (>75% of cases)

-   Upper extremity Doppler of CTA

Surgical Considerations 

-   Given course, can be extensively mobilized and repaired in
    end-to-end fashion in 50% of cases. Otherwise, treatment with an
    interposition graft

**What is the mechanism and management of upper extremity radial/ulnar
artery trauma?**

Mechanism and Pattern

-   Associated with significant soft tissue pattern

Diagnostic Considerations

-   Pulse deficit in >80% of patients 

-   Doppler based Allen test-confirm radial/ulnar contribution to palmar
    arch

Surgical Considerations 

-   If Allen test reveals a patent palmar arch, the injured artery can
    be ligated 

-   If palmar arch is not patent in the absence of contribution of the
    injured artery, it should be repaired

-   If both are damaged, preference to ulnar artery as dominant
    contribution to hand 

-   Generally, repair can be done in an end to end fashion given
    mobility of the vessel

## Compression Syndromes

**The** **main syndromes are quadrilateral space syndrome and humeral
compression of the axillary artery. What important information here do
our listeners need to know?**

### Quadrilateral space syndrome

Anatomy:

-   Bordered by teres minor superiorly, humeral shaft laterally, and
    teres major inferiorly, and long head of triceps muscle medially

-   Posterior humeral circumflex artery and axillary nerve in space

Pathophysiology

-   Compression of posterior humeral circumflex occurs with
    abduction/external rotation 

-   Typically seen with chronic overhand motion athletes
    (pitchers/volleyball players)

-   Vascular-repetitive mechanical trauma to posterior circumflex
    humeral artery 

-   Neurogenic-fixed structural impaction of quadrilateral space by
    fibrous bands or space-occupying lesions

Presentation

-   Muscle atrophy, paresthesias, poorly localized shoulder pain and
    pain in quadrilateral space

Treatment

-   Medical: Oral anti-inflammatory medications, PT, limitation of
    activities

-   Surgery: decompression with neurolysis/excision of fibrous bands or
    other space occupying lesions 

### Humeral head compression of axillary artery

Anatomy:

-   3rd portion of axillary artery compressed by head of humerus

Etiology/Pathophysiology:

-   Arm is abducted and externally rotated with downward compression of
    humeral head to axillary artery

Presentation:

-   Arm fatigue, loss of pitch velocity, finger numbness, Raynaud,
    cutaneous embolization

Diagnosis:

-   Provocative maneuvers with impedance of flow through axillary artery
    on ultrasonography

-   Arteriography with rest and provocative position

Treatment:

-   Supportive with avoidance of throwing motion

-   Surgical-saphenous vein patch for no improvement or structural
    injury may require resection with saphenous vein bypass anatomically
    or extra-anatomic tunneling above pec minor

## Thoracic Outlet Syndrome

**27 Nov 2019:** *Nedal Katib, Prince of Wales, Sydney Australia*

```{r echo=FALSE}
library(embedr)

embed_audio("https://traffic.libsyn.com/audiblebleeding/VSITE_TOS_02auph.mp3")

```

Thoracic Outlet Syndrome = A constellation of signs and symptoms
relating to the compression of the neurovascular structures that occurs
as these structures travel between the thoracic aperture and the upper
limb.

Types: Neurogenic, Venous and Arterial 

-   vTOS -- 2-3%

-   aTOS -- 1%

-   nTOS --  >95% [@humphries124ThoracicOutlet2019]

### Anatomy

Understanding the anatomy of what is collectively referred to as the
thoracic outlet is the best way to thoroughly appreciate this topic.

Three spaces where the neurovascular structures are at risk of
compression:

1.  Interscalene Triangle 

2.  Costoclavicular Passage [@garygwindAnatomicExposuresVascular2013]

3.  Subcoracoid Space [@garygwindAnatomicExposuresVascular2013]

**Interscalene Triangle:**

Appreciating the attachments of the Anterior and Middle Scalene Muscles
on the first rib becomes important in the diagnosis of the various types
and also the ultimate surgical management of the compression.

**Anterior Scalene:**

Attachments: Anterior Tubercles of the four 'typical' cervical vertebrae
(3-6) AND the scalene tubercle on the upper surface of the first rib.

**Middle Scalene:**

Attachments: The posterior tubercles and intertubercular lamellae of all
the cervical vertebrae AND the Quadrangular area between the neck and
subclavian groove of the first rib. [@mcminnLastAnatomyRegional2019]

**The First Rib:** 

-   The broadest and flattest of the ribs and is an 'Atypical Rib'. 

-   The upper surface of the first rib has the scalene and quadrangular
    tubercles for attachments of the anterior and middle scalene muscles
    respectively. There are also three grooves for the Subclavian Vein,
    artery and the Lower Trunk of the Brachial Plexus. 

-   The Inferior Surface is smooth and inferior and medially has an
    attachment for the suprapleural membrane, Sibson's fascia AKA
    scalenus minimus, which is tethered to the C7 vertebrae. 

-   This is the passage of the subclavian vein largely as it emerges
    through the tight space created by the clavicle, the subclavius
    muscle and the costoclavicular ligament and also more posteriorly
    this can also compress the artery and nerves as the space can also
    be narrow in relation with the scapula and subscapularis.
    [@garygwindAnatomicExposuresVascular2013]

**Subclavius Muscle:**

-   Attached to the costochondral junction of the first rib and is
    inserted into the subclavian groove on the inferior surface of the
    clavicle. [@mcminnLastAnatomyRegional2019]

-   This space is best appreciated by intimate knowledge of three
    things:

    -   The Coracoid Process and its attachments

    -   The Pectoralis Minor Muscle

    -   The Clavipectoral Fascia

**The Coracoid Process:** 

-   Arising from the Scapula as a 'process', this broad-based bony
    landmark offers attachment to muscles and ligaments.

-   The relevant attachments being the pectoralis minor muscle occupying
    the medial border for about 2cm behind its tip. The tip itself
    having a medial and lateral facet for the short head of biceps and
    the coracobrachialis muscles respectively.

**Pectoralis Minor Muscle:**

Attached to the bone of the third, fourth and fifth ribs AND the medial
border of the coracoid process. 

**Clavipectoral Fascia:**

A sheet of fascial membrane that fills the space between the clavicle
and pectoralis minor splitting and encompassing the subclavius muscle.
Its superior portion is what can be thickened and become a tight band
referred to as the costocoracoid ligament. 

**Phrenic Nerve Anomaly:**

The Phrenic Nerve normally runs anterior to the Subclavian Vein. A rare
anomaly is the nerve compressing the vein anteriorly and in very rare
circumstances due to the timing of development can run through the vein
itself.

Anomalous anatomy can also cause TOS especially when patients have a
Cervical Rib and anomalous first ribs or a congenital band attaching to
the first rib.

-   Incidence of anomalous first ribs and cervical ribs is 0.76% and
    0.75% respectively. 

-   Incidence of bands are as high as 63% in the general population.
    [@humphries124ThoracicOutlet2019]

nTOS  

-   Scalene Triangle compression -- most common cause of brachial plexus
    and neurogenic TOS

-   Cervical Rib and Anomalous First Rib

aTOS

-   Cervical Rib and Anomalous First Rib

-   Scalene Triangle compression

vTOS

-   Costoclavicular Passage

-   Subcoracoid Space

### Diagnosis and Evaluation

#### Patient History

-   Identify symptoms and thoroughly interrogate timing

-   Exclude history of trauma

-   Associated symptoms like headache, visual disturbance, neurology in
    the upper limb

-   Exclude Carpal Tunnel and Antecubital Tunnel Syndromes if symptoms
    are isolated to the arm or forearm or hand

-   Patients with vTOS may present acutely and have acute or subacute
    Upper Limb DVT

-   Patients with aTOS need to be investigated and assessed urgently
    given risk of ischemia. 

#### Clinical Examination

Provocative maneuvers are largely used for nTOS. While these are
described and mentioned in most texts their utility largely is beyond
the scope of a vascular surgeon's assessment and diagnosis of nTOS.

**Adson Test**

-   Extended abducted and externally rotated arm -- palpate radial pulse

-   Rotate and laterally flex the neck to the ipsilateral side while
    inhaling deeply.

-   A positive test results in reduction or complete obliteration of
    radial pulse

**Roos Test / EAST test**

-   Patient seated and both arms abducted 90 degrees and externally
    rotated and elbows flexed at 90 degrees. 

-   Open and close hands for 3 minutes or until pain or paraesthesia
    sets in.

**Elveys Test**

-   Abduct both arms to 90 degrees with elbows extended and dorsiflex
    both wrists.

-   If pain is elicited as wrists dorsiflexed then test is positive.

-   A further manoeuvre is then performed, laterally flex the head on
    each side, if pain is elicited on the contralateral side to which
    the head is flexed then test is positive.
    [@humphries124ThoracicOutlet2019]

#### Non-invasive imaging or vascular lab studies

-   DBI

-   Arterial Duplex

-   Venous Duplex 

-   CT -- CTV commonly performed in acute upper limb DVT and suspicion
    of vTOS

-   CTA for the evaluation of aTOS and excluding other causes of
    embolisation

-   MRI -- for further evaluation of the anatomy and related
    neurovascular compression

-   Electromyography and Nerve Conduction Studies for nTOS

**Paget Schroetter Syndrome**

-   First defined by Hughes in 1949 in reference to Sir James Paget who
    in a hundred years earlier defined acute arm swelling and pain as
    possibly related to vasospasm and then von Schroetter who in 1884
    attributed to the presentation to subclavian and axillary vein
    thrombosis. [@humphries123ThoracicOutlet2019]

-   Now vTOS and Paget Schroetter Syndrome are used synonymously.

-   Paget Schroetter Syndrome accounts for 10-20% of all upper extremity
    deep vein thrombosis. [@sekharYearbookVascularEndovascular2018]

### Rib Resection approaches

+------------+------------+------------+------------+
| Type       | Tr         | Supr       | Infr       |
|            | a          | a          | a          |
|            | nsaxillary | clavicular | clavicular |
+============+============+============+============+
| Advantages | C          | -Good for  | -Good      |
|            | o          | scalene    | access for |
|            | smetically | triangle   | venous d e |
|            | more       | access and | c          |
|            | appealing  | debulking  | ompression |
|            | as it has  | and        |            |
|            | a limited  | cervical   | -Allows    |
|            | hidden     | rib        | for        |
|            | scar       | resection  | excision   |
|            |            |            | of         |
|            |            | -Required  | subclavius |
|            |            | for aTOS   | muscle and |
|            |            | if         | c o s t    |
|            |            | arterial r | o          |
|            |            | e c        | clavicular |
|            |            | o          | ligament   |
|            |            | nstruction |            |
|            |            | necessary  |            |
+------------+------------+------------+------------+
| D i        | -Difficult | -Unable to | -Unable to |
| s          | to         | decompress | expose     |
| advantages | visualise  | venous     | subclavian |
|            | the        | c          | artery     |
|            | anatomy,   | ompression |            |
|            | dependent  | or         | -Difficult |
|            | on good    | visualise  | to access  |
|            | assistance | vein       | most       |
|            |            | adequately | posterior  |
|            | -Risk of   |            | aspect of  |
|            | T1 nerve   | -   c      | rib        |
|            | root       |     o      |            |
|            | injury,    | smetically | \- C       |
|            | subclavian |     less   | o          |
|            | vein       |            | smetically |
|            | injury and | appealing  | less       |
|            | arterial   |            | appealing  |
|            | injury     |            |            |
|            | with       |            |            |
|            | limited    |            |            |
|            |            |            |            |
|            | exposure   |            |            |
|            | to repair  |            |            |
|            |            |            |            |
|            | -Not able  |            |            |
|            | to         |            |            |
|            | approach   |            |            |
|            | cervical   |            |            |
|            | ribs,      |            |            |
|            | scalene    |            |            |
|            | triangle   |            |            |
|            | or patch   |            |            |
|            | vein.      |            |            |
+------------+------------+------------+------------+

### vTOS

#### Presentation

-   Incidence: 2/100,000 persons

-   Age: 18 years to 30 years
    [@illigComprehensiveReviewPagetSchroetter2010]

-   M>F

    -   Upper Limb edema, pain and cyanosis. Edema affects the shoulder,
        arm and hand and characteristically non pitting.

    -   Collateral vein dilatation over the shoulder, neck and anterior
        chest wall to accommodate for the increased venous hypertension.
        [@humphries123ThoracicOutlet2019]

    -   Pain on exertion of the upper limb described as stabbing, aching
        or tightness.

    -   The reported incidence of PE following Upper Limb DVT is \<12%.
        [@humphries123ThoracicOutlet2019]

#### History

-   A differential diagnosis for Upper Limb DVT

    -   vTOS

    -   Congenital Phrenic Nerve anomaly

    -   History of Fracture, Clavicular Fracture and malunion

    -   Repetitive arm provocative manoeuvres, check occupation and
        history of body-building

        -   Pectoralis Minor Hypertrophy.

-   Exclude Pulmonary Embolism 

-   Exclude Venous Gangrene and Phlegmasia of the upper limb

#### Goals of therapy for vTOS

Limited evidence due to lack of RCT's. Majority of evidence based on
retrospective studies. 

-   Prevent immediate risk 

-   Return patient to unrestricted use of the affected extremity

-   Prevent recurrence of thrombosis without the need of long-term
    anticoagulation

-   Prevent long term Post Phlebitic Limb Syndrome

#### Initial management strategy

-   Initial Management

-   As per ACCP Guidelines: Initial management is anticoagulation
    regardless of etiology. [@kearonAntithromboticTherapyVTE2016]

    -   The limitations of anticoagulation alone are that the slow
        recanalization of the thrombus may lead to eventual valvular
        damage and intravenous scarring.
        [@sekharYearbookVascularEndovascular2018]

    -   Thrombolysis has been considered superior to anticoagulation
        alone in minimizing valvular damage due to residual clot.
        [@urschelSurgeryRemainsMost2008]

    -   Systemic Lysis -- non favoured due to risk of intracranial
        haemorrhage.
        [@grunwaldCatheterDirectedThrombolysisTreatment2004]

    -   Catheter Directed Lysis (CDT) -- carries a lower risk of
        intracranial haemorrhage.

-   Optimal timing of CDT

    -   Within 14 days of onset of thrombosis. Excellent results have
        been reported following CDT if initiated before 14 days.
        [@wilsonFibrinolyticTherapyIdiopathic1990]

-   Surgical indications for vTOS

    -   After initial management patients are generally divided into two
        groups, unsuccessful or successful thrombolysis.

    -   Persistent stenosis or signs of extrinsic compression, on
        venography, has generally been perceived as a significant risk
        of recurrent thrombosis.

    -   Surgery for vTOS remains to be mainly Rib Resection and
        decompression of the subclavian vein with or without venolysis
        and patch plasty either surgical or endovenous.

#### Controversy around vTOS

-   There is a lack of consensus around the necessity of surgical rib
    resection, the timing and the requirement for vein patch plasty.

-   Options post recanalization:

    -   Deferring surgical decompression for 1-3 months after
        thrombolysis to allow for healing of the venous endothelium and
        resolution of the acute inflammatory process.
        [@humphries123ThoracicOutlet2019]

    -   Decompression during the same admission, as the thrombolysis,
        with the main benefit being to reduce the risk of re-occlusion.
        [@humphries123ThoracicOutlet2019]

#### Landmark papers regarding vTOS and what are the take home messages

1.  Lugo J et al -- Acute Paget Schroetter syndrome: does the first rib
    routinely need to be removed after thrombolysis? Annals of Vascular
    Surgery 2015 [@lugoAcutePagetSchroetter2015]

    1.  Systematic literature review analysis. Patients divided into
        three groups

        1.  First Rib resection (FRR) -- n=448

        2.  First Rib resection and endovenous venoplasty (FRR and
            PLASTY) n=68

        3.  No further intervention after Thrombolysis n=168

    2.  Symptom relief after initial follow up more likely in FRR (95%)
        and FRR and PLASTY (93%) compared to no rib removed (54%) --
        p\<0.0001

    3.  Results showed superior patency with FRR and PLASTY and FRR
        compared to anticoagulation alone. 

    4.  Conclusion was that patients are more likely to experience
        greater long-term results with FRR compared to no FRR. 

2.  Sajid MS et al -- Upper limb vein thrombosis: a literature review to
    streamline the protocol for management. Acta Haematology 2007
    [@sajidUpperLimbDeep2007]

    1.  a comprehensive review identifying the key papers on this topic
        and allows for a clear view of the best management strategy. 

### aTOS

#### Presentation

-   Most common: Hand ischemia due to arterial compression or
    microembolization

-   Exertional pain 

-   Unilateral Raynaud's Phenomena

-   Subclavian artery aneurysm and pulsatile supraclavicular mass

-   Rarely retrograde embolisation and neurological symptoms 

-   Clinical Examination

    -   Audible Bruit / Palpable thrill over the supraclavicular fossa

    -   Pulsatile mass

    -   Distal ischemic lesions in the distal hand -- Splinter
        haemorrhages

    -   Positive Adson Test

#### Important differentials for aTOS

-   Trauma

-   Primary and Secondary Raynaud's Phenomena

-   Small Vessel Vasculitis

-   Connective Tissue Disorders

-   Thromboangiitis Obliterans

-   Arterial Embolisation -- Aortic or Central Source

-   Radiation Arteritis

-   Atherosclerotic / Dissection causes

#### Most common causes of aTOS - The different Anatomical Abnormalities causing aTOS (incidence %)

-   Cervical Rib (60%)

-   Anomalous First Rib (18%)

-   Fibrocartilaginous band (15%)

-   Clavicular Fracture (6%)

-   Enlarged C7 transverse process (1%) [@boll122ThoracicOutlet2019]

#### Scher Staging of aTOS

-   Stage 0: Asymptomatic

-   Stage 1: Stenosis of Subclavian Artery with minor post stenotic
    dilatation with no intimal disruption

-   Stage 2: Subclavian artery aneurysm with intimal damage and mural
    thrombus

-   Stage 3: Distal embolisation from subclavian artery disease

#### Management considerations with aTOS

-   Symptomatic patients are generally indicated for treatment. Unlike
    asymptomatic patients in whom it may be appropriate to manage
    conservatively. [@boll122ThoracicOutlet2019]

-   Supraclavicular rib resection is the most suitable for adequate
    arterial reconstruction. Transaxillary has been argued to offer more
    complete rib resection however arterial repair is not possible in
    this approach.

-   Subclavian artery repair is necessary in Scher Stages 2 and 3 and in
    some cases 1. Arterial repair with conduit either GSV, Femoral Vein
    or prosthetic have been described. Ringed PTFE offers good patency
    and resistance to kinking in this functional anatomical location. 

### nTOS

#### Demographics

Neurogenic TOS is largely a clinical diagnosis with symptoms and signs
pertaining to nerve compression most commonly the lower trunk of the
brachial plexus. 

-   F>M -- 70% Female

-   Ages 20-40

-   Occupational Exposure

-   Trauma history

#### Presentation of nTOS

-   Symptoms

    -   Pain and Paraesthesia

    -   Numbness

    -   Neck pain

    -   Headaches

    -   Weakness

    -   Swelling

-   Positional Effects

    -   Reproducible exacerbation of symptoms

    -   Lying supine with arms overhead

    -   Overhead activities -occupational or recreational

-   Weakness and Muscle Atrophy

    -   Hypothenar atrophy

    -   Dropoff in athletic performance

    -   Inability to carry out activities of daily living 

#### The role of the Vascular Surgeon with nTOS

Often these patients have already seen multiple specialists and
physiotherapists.

-   Exclude other causes

-   Confirm diagnosis -- Neurophysiologic Tests (EMG and NCS)

-   Seek alternate opinion

-   Trial of Physiotherapy and non-operative management

-   Be selective in patients who may require surgery

Surgery with Rib resection often is accomplished with transaxillary or
supraclavicular approach, particularly if scalenectomy or cervical rib
resection is necessary.
