---
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

# Upper Extremity and Thoracic Outlet

**21 Jan 2021**: *Kush Sharma, MD and Ashraf Mansour, MD*

```{r echo=FALSE}

library(embedr)

embed_audio("https://traffic.libsyn.com/audiblebleeding/VSITE_Upper_Extremity_post_auphonic.mp3")

```

## Anatomy/ Exposure of Vessels

**What are the zones of the upper extremity?**
[@illig57UpperExtremity2019; @illig57UpperExtremity2019]

Division of the upper extremity into three zones:
<!--# Would be great to link to an free open source anatomy text -->

1.  Intrathoracic Zone: Inclusive of the aortic arch and its branches,
    innominate artery (dividing into right subclavian and right common
    carotid), left subclavian artery and left common carotid artery, as
    well as the innominate veins, and SVC.

    -   Arch classification helps determine difficulty of accessing the
        right innominate from a femoral approach. Type 1 innominate does
        not break the plane of the outer curve, Type 2 it is between the
        outer curve and inner curve and Type 3 it falls below the inner
        curve of the aortic arch.

2.  Thoracic Outlet: Extends from the base of neck to the axilla.
    Includes the subclavian, proximal vertebral, and proximal axillary
    arteries, as well as the corresponding veins).

3.  Axilla to Fingers: The arm.

**What are some common exposures for major upper extremity arteries?**

1.  Right Subclavian Artery: For proximal artery exposure a medial
    sternotomy is required, for the mid-distal artery a right
    supraclavicular approach is needed.

2.  Left Subclavian Artery: An anterolateral thoracotomy is required in
    emergent settings for proximal left subclavian artery control. For
    proximal, mid and distal exposure is required a "trap door" exposure
    is used: anterolateral thoracotomy, third space sternotomy, and left
    supraclavicular incision.

3.  Supraclavicular Incision: After division of the platysma and
    clavicular head of the SCM, a fat pat of varying thickness is
    encountered which contains the omohyoid muscle. This should be
    divided and placed supero-laterally. At this point, the anterior
    scalene muscle is exposed medially with the phrenic nerve running in
    a lateral to medial direction, which should be identified and
    preserved. In a carotid/subclavian bypass division of the anterior
    scalene should be performed as close to the first rib as possible.
    Once divided the subclavian artery is exposed.

4.  Axillary Artery: Exposure is gained through an infraclavicular
    incision, below the middle third of the clavicle. Pectoralis major
    is divided and pectoralis minor is freed at the lateral margin of
    the wound. The axillary vein should be identified and followed
    supero-deeply to identify and expose the artery.

Anatomically bound by the first rib proximally and the lateral edge of
the teres major muscle distally. For exposure of the first part of the
axillary artery, the ipsilateral arm is abducted approximately 90
degrees and a horizontal skin incision 2 cm below the middle third of
the clavicle should be made. The underlying pectoralis major is split by
bluntly separating the fibers, exposing the tough clavipectoral fascia.
At the lateral wound margin, the pectoralis minor can be freed and
laterally retracted. The axillary vein is the first structure to be
encountered in the fascia and the artery lies just superior and deep to
the vein. Make sure to protect the nerves of the brachial plexus which
lie deep to the first part of axillary artery and are at risk of injury
during blind placement of occluding arterial clamps.
[@garygwindAnatomicExposuresVascular2013]

**What steps are involved for brachial, radial and ulnar artery
exposure?**

**Brachial Artery, Upper Arm:** The brachial artery's superficial
location makes it vulnerable to injury and accounts for most vascular
injuries of the upper extremities. Brachial artery exposure involves a
5-8 cm longitudinal incision in the groove between the biceps/triceps
muscles on the medial aspect of the arm. In the lower half of the arm,
take care to avoid basilic vein damage as it runs in the subcutaneous
tissue. The neurovascular bundle is exposed by incising the deep fascia
at the medial border of the biceps muscle, which then should be
retracted anteriorly. The basilic vein should be indetified and
retracted into the posterior wound. On opening the brachial sheath the
median nerve is the most superficial structure and should be identified,
protected and retracted. The brachial artery lies deep to the median
nerve and is flanked by two brachial veins. Be midful that the ulnar
nerve lies posteriorly.

**Brachial Artery, Antecubital Fossa:** Exposure of the brachial artery
in the antecubital fossa requires a transverse skin incision 1 cm distal
to the midpoint of the antecubital crease. After deepening, mobilize the
basilic vein medially to avoid injury. The medial antebrachial cutaneous
nerve should be identified and protected. The bicipital aponeurosis
should be divided to exposure the brachial artery, which is flanked by
two deep veins crossing tributaries. To isolate the brachial artery
ligation and division of these tributaries is required.

The brachial artery bifurcates at the radial tuberosity into its radial
and ulnar branches. Immediately after the ulnar artery arises from the
bifurcation origin, it gives off a short common interosseous branch,
which bifurcates at the hiatus in the proximal interosseous membrane.

**Radial Artery at the Wrist:** This requires a 2-3 cm longitudinal
incision generally between the palpable radial artery and cephalic vein.
Incision of the antebrachial fascia medially to the radius should expose
the radial artery. Two veins accompany the artery and should be
dissected away to isolate the artery. The superficial radial nerve and
its medial/lateral branches course between the cephalic vein and radial
artery in this area.

**Ulnar Artery:** The ulnar artery courses beneath the superficial
flexor muscles of the proximal forearm, emerging near the ulnar border
at the midpoint between the elbow and the wrist. In the distal forearm,
the ulnar artery runs just deep to the antebrachial fascia and is easily
exposed through a longitudinal incision placed radially to the flexor
carpi ulnaris. The palmar branch of the ulnar nerve courses
superficially to the antebrachial fascia and should be preserved during
arterial exposure to avoid injury.

**What common aberrant upper extremity/arch anatomy is important to be
aware of?**

-   **Bovine Arch:** Where the innominate and left common carotid
    arteries share a common origin.[@layton2006]

-   **Aberrant Vertebral Artery:** Where the vertebral artery arises
    directly from the aortic arch.

-   **Aberrant Right Subclavian Artery:** Where there is no innominate
    artery and instead the right common carotid arises directly from the
    aorta. Confusingly the right subclavian artery, then originates from
    the aortic arch distal to the left subclavian artery. To supply the
    right arm it therefore, needs to pass behind the esophagus.
    (Dysphagia Lusoria)

-   Thoracic aortic disease is associated with aberrant anatomy, with
    high rates of bovine arch (25%), isolated L vertebral (6-8%), and
    aberrant R subclavian (1-2%).[@dumfarth2015]

## Epidemiology, Etiology, and Diagnostic Evaluation

**How does evaluation of upper extremity ischemia differentiate from
lower extremity ischemia?** [@shuja117UpperExtremity]

-   Upper extremity ischemia \<5% of patients with limb ischemia and in
    contrast to lower extremity, atherosclerosis is not a major
    contributor to upper extremity ischemia

-   Vast majority of cases caused by autoimmune/connective tissue
    disorders

**How can upper extremity disease be classified?**

Anatomic Location:

-   Large vs. Small Vessel

Disease Process:

-   Vasospastic or occlusive. Vasospastic disease is more responsive to
    pharmacologic management while occlusive requiring
    endovascular/surgical management.

**How should patients be evaluated who have concern for upper extremity
disease?**

Diagnostic Evaluation

1.  Detailed H+P evaluation (pulse palpation, auscultation at
    supraclavicular/infraclavicular fossa may reaveal a bruit concerning
    for subclavian artery stenosis, upper extremity neurovascular/skin
    exam)

2.  Brachial/forearm blood pressures and if suspected claudication,
    measured at rest and 2-5 minutes after exercise. Look for a gradient
    of \>20 mmHg is considered significant

3.  Some or all of 6 P's of acute limb ischemia with symptoms occurring
    within 14 days are deemed acute

4.  Doppler insonation of radial, ulnar, palmar, and digital arteries

5.  Vascular Lab Evaluation

    1.  Segmental Pressure Measurements

    2.  Duplex Ultrasound (look for large vessel occlusive disease)

6.  Other Imaging

    1.  CTA/MRA

7.  Clinical Lab tests

    1.  Inflammatory disorders-CBC, ESR, ANA, RF

    2.  Hypercoagulable screening

## Operations/Procedures

**What is the procedure of choice in acute limb ischemia of the upper
extremity?**

With acute presentation of upper limb ischemia and a localizing
examination, may be reasonable to proceed directly to OR for embolectomy
to minimize ischemia time.[@wahlberg2006; @henke2009]

**What are some postoperative complications of an embolectomy for acute
limb ischemia?**

Brachial sheath hematoma - parasthesias and weakness in the median nerve
distribution require emergent re-exploration and decompression. Even
small hematomas that are not readily identifiable on clinical exam, can
cause compression on the median nerve and are a surgical
emergency.[@tran2011]

Compartment Syndrome - If prolonged ischemia or no hematoma found, and
forearm compartments appear tense, then should proceed with fasciotomies
to treat forearm compartment syndrome. Rarely performed prophylactically
due to significant morbidity.[@gelberman1981; @kistler2018;
@leversedge2011]

-   Three compartments - forearm volar (flexor, superficial, and deep),
    mobile wad (lateral), and dorsal (extensor, superficial, and deep).
    Volar compartments most susceptible to ischemia and compartment
    syndrome, most vulnerable muscle are flexor digitorum profundas and
    flexor pollicus longus.[@ronel2004]

-   Volar/Henry approach decompresses lateral and volar compartments
    with a single incision, includes carpal tunnel release.

**What are some indications for carotid-subclavian bypass?**

1.  Atherosclerosis

2.  Staged revascularization prior to TEVAR for aneurysmal disease
    requiring coverage of the LSA

**How does the exposure differentiate in transposition vs bypass?**

Exposure (Transposition vs Bypass)

-   Arterial transposition via a short, transverse cervical incision
    above the clavicle between two heads of SCM (bypass is lateral to
    entire SCM)

-   Sub-platysmal flaps created and avoid EJ vein damage

-   Omohypoid divided between heads of SCM and IJ mobilized laterally
    (bypass IJ is mobilized medially to expose CCA and care must be
    taken to avoid phrenic nerve in more lateral approach)

-   CCA is reflected medially with vagus nerve

-   On the left side, the thoracic duct is identifiable and divided
    followed by dividing the vertebral vein

-   Subclavian artery and proximal branches identified (anterior scalene
    is in lateral dissection)

**What are some common complications after carotid subclavian bypass in
order of highest to lowest incidence?**

Complications [@voigtOutcomesCarotidsubclavianBypass2019]

1.  Phrenic nerve palsy (most common) - most often managed
    conservatively.

2.  Recurrent laryngeal palsy

3.  Lymphatic leak

4.  Neck hematoma

**When carotid-subclavian bypass compared to transposition?**

1.  Vertebral artery takes origin from the subclavian artery in a very
    proximal position or is dominant over the contralateral side, then
    bypass preferred. [@moraschTechniqueSubclavianCarotid2009d]

2.  For coronary-subclavian steal with patent internal mammary artery to
    coronary artery bypass graft, then Bypass (a carotid-subclavian
    transposition requires a more proximal clamp with occlusion of
    inline antegrade flow to the coronary bypass during the procedure)
    [@cuaReviewCoronarySubclavian2017]

## Vaso-occlusive disease

**What are causes and symptoms associated with subclavian/axillary
occlusive disease?** [@jacklcronenwettVascularDecisionMaking2020]

-   Etiology: Atherosclerosis is the most common cause of
    subclavian/axillary occlusive disease. Left SCA \> Right
    involvement. Less common causes include Takayasu disease, giant cell
    arteritis, or arterial TOS

-   Symptoms: Upper extremity arm/hand ischemia or neurologic symptoms
    due to subclavian-vertebral steal. Because significant collaterals,
    minimal pain on exertion even with subclavian occlusion

-   Innominate disease can present with complex steal syndrome from the
    vertebral artery.[@rodriguez2016]

**What are causes and symptoms associated with brachial/forearm
occlusive disease?**

-   Etiology: MCC of brachial artery occlusion is cardiac origin
    embolus. Atherosclerosis RARELY affects the brachial artery. Distal
    axillary/proximal brachial stenosis can be from repetitive trauma
    from crutch use.

    -   Radiation arteritis - may be seen with radiation after breast
        cancer therapy. Lesions characteristically tapered with smooth
        border. May occur alongside a brachial
        plexopathy.[@goldstein2010; @modrall2003]

-   Forearm occlusive disease can be seen in advanced ESRD/DM where
    calcific atherosclerosis of radial/ulnar arteries is present. Less
    common causes include Buerger'as disease or Raynaud Phenomenon

**How/when is upper extremity occlusive disease treated?**

-   Indications for treatment

    -   70% stenosis and associated symptoms or deficit

    -   Lesion with ipsilateral ICA requiring treatment

    -   Pre-occlusive lesion in a good surgical candidate (\>5y life
        expectancy)

-   Innominate or SCA Occlusive Disease

    -   Endovascular with balloon expandable stent via femoral or
        ipsilateral brachial artery.
        [@chatterjeeAngioplastyAloneAngioplasty2013;
        @bradaricEndovascularTherapyStenoOcclusive2015;
        @sahaSubclavianArteryDisease2017; @palchik2008; @mordasini2011]
        Preferred in:

        -   Short segment or ostial disease with adequate distance to
            the vertebral artery origin.

        -   History of neck surgery or radiation.

    -   Surgery:

        -   Bypass from aortic arch through median sternotomy - Most
            durable option, particularly for dense R innominate disease,
            so preferred in patients who can tolerate. Endarterectomy
            not appropriate if disease involves the origin.[@aiello2011;
            @mansukhani2018; @berguer1998; @daniel2014; @reul1991;
            @berguer1999; @byrne2007]

            -   Complications include compression of the graft on
                sternal closure, so proximal anastomosis should be
                placed to the side of the aorta.

        -   Ipsilateral CCA to subclavian artery (bypass or
            transposition) - Recent NSQIP study showed no difference in
            stroke and death for CS bypass vs transposition, therefore
            procedure should be determined by anatomic
            constraints.[@cinà2002; @vandervliet1995] Small studies and
            systematic reviews have shown that patency of carotid
            subclavian bypass is better with synthetic graft than with
            autologous vein.[@aburahma2000; @illuminati2018;
            @ziomek1986]

        -   Contralateral CCA (anterior or retropharyngeal)

-   Brachial/forearm Occlusive disease

    -   Endovascular: PTA evidence is anecdotal with stents for lesions
        unresponsive to PTA or dissection following
        angioplasty.[@cheun2019; @nasser2014; @dineen2007]

    -   Surgery: GSV vein bypass remains standard for revascularization
        with bypasses to superficial or deep palmar arch have good
        patency rates. Tunneling is subcutaneous if to distal ulnar or
        superficial palmar arch whereas anatomical to distal radial
        artery over the anatomic snuffbox.[@chang2003; @masden2012;
        @spinelli2010]

## Vasospastic Disorders

### Iatrogenic or Vasopressor Induced Ischemia

**You are called to the intensive care unit for a septic patient with
acute bilateral upper limb ischemia, what is the management strategy?**

Critically ill patients on vasopressors with bilateral upper extremity
ischemia are often managed best with supportive measures and attempts to
wean vasopressors (especially norepinephrine).[@landry2018]

Limited role for radial artery embolectomy, only in clear ischemia
isolated to the distribution of an occluded radial artery, often
secondary to line placement.[@valentine2005]

### Raynaud's Phenomenon

**What is Raynaud's and what causes it?** [@shuja117UpperExtremity;
@landry141RaynaudPhenomenon2019]

-   Exaggeration of normal physiologic response with episodic pallor or
    cyanosis of the fingers caused by small digital artery
    vasoconstriction occurring in response to cold or emotional stress.
    There is an abnormality with sympathetic nervous system, resulting
    in a multifactorial problem involving a combination of vascular,
    neural, and humoral factors.

**What are the subtypes of Raynaud's phenomenon and what is the
underlying pathology?**

-   Primary: Raynaud's disease-idiopathic form that is a benign process
    not associated with structural vascular change. Triggers include
    (cold, emotional stress, caffeine) resulting in digital smooth
    muscle contraction and temporary digital hypoperfusion.

-   Secondary: Fixed vascular obstruction to blood flow decreasing
    threshold for cold induced vasospasm or progress to tissue loss.
    Diseases associated include mixed connective tissue disease, SLE,
    and rheumatoid arthritis, and scleroderma (accounts for 80-90% of
    cases). In setting of lower digital blood pressure, symptomatic
    digital ischemia or tissue loss under low stress conditions. With
    cold/emotional stress, vasoconstrictive response of digital artery
    smooth muscle further causes arterial closure and resultant symptoms

**What are diagnostic criteria for Raynaud's?**

-   Clinical (Progression of ischemia with white -\> blue -\> red finger
    discoloration. Episodes can be self-limited and may last from less
    than a minute, but generally not longer than 10-20 minutes

-   Qualitative testing for severity of cold sensitivity in Raynaud's
    syndrome can be useful. Most basic test is cold sensitivity and
    recovery after ice water immersion. \>10 minutes return to baseline
    pressure concerning for Raynaud's

-   Segmental pressures with finger systolic blood pressure can
    differentiate purely vasospastic vs occlusive disease. Difference of
    more than 15 mm Hg between fingers or absolute finger pressure \<70
    mm Hg may indicate occlusive disease

-   Serologic evaluation (ANA/RF)

**What are appropriate treatments for Raynaud's phenomenon?**

1.  Medical-cold/tobacco avoidance lead to improvement in around half of
    patients. Calcium channel blocker (nifedipine ER 30mg qday) and
    losartan (50mg BID) have been shown to be most
    effective.[@landry2013; @wigley2016] Other drugs include alpha
    blocker, sildenafil, fluoxetine (SSRI), reserpine, cilostazol,
    captopril.

    1.  NOT GOOD OUTCOMES IN PATIENTS WITH ARTERIAL OBSTRUCTION. If
        there is an asymmetric vascular examination, then further
        non-invasive vascular imaging is needed.

2.  Surgical-thoracic sympathectomy (used for treatment of digital
    artery vasospasm/digital ischemic ulceration). For vasospasm,
    thoracic sympathetcomy is initially successful, but symptoms return
    generally within 3-6 months.

3.  Immunosuppression/immunomodulation for connective tissue disorders
    associated with secondary Raynaud phenomenon

See Landry's Review in JVS for a good treatment algorithm.[@landry2013]

### Ergotism

**What is Ergotism?** [@jamescstanleyCurrentTherapyVascular2014]

-   Etiology: Ergot is a parasitic fungal disease that has a particular
    prevalence for infecting rye plants and ergot alkaloids have been
    linked to epidemic poisonings that manifested as ergotism from
    consumption of rye

-   Modern day is rare

**What causes Ergotism and how do patients present?**

-   Ergotamine is chemically like endogenous catecholamines/indolamines
    and when applied clinically, it behaves as an agonist to
    alpha-adrenergic, sertoninergic, and dopaminergic receptors. Despite
    limited bioavailability, vasocontrictive effects have been reported
    to last for 24 hours or longer

-   Gangrenous-mild limb pain followed by burning pain/shooting and

-   Convulsive-heaviness in limbs and head associated with diarrhea.
    Could result in tonic-clonic spasms

**How can you diagnose Ergotism and what is the process for treating
this disease?**

Upper extremity ischemia (i.e. digital ulceration) in the setting of
ergot alkaloid use (typically for migraines)

Treatment:

-   Volume expansion and IV heparin as anticoagulation

-   IV infusion of nitroprusside, nitroglycerin, ilioprost or
    combination

-   Infusion of Ca 2+ channel blockers

-   Surgical: for thrombosis, consider thrombolysis

### Buerger's Disease

**How is Buerger's disease categorized?**
[@jacklcronenwettVascularDecisionMaking2020; @lejoncour2018]

-   Non-atherosclerotic, segmental, inflammatory disease of small/medium
    sized arteries in distal extremities of tobacco users distinct from
    either atherosclerosis of immune arteritis

**What clinical criteria can help diagnose Buerger's?**

-   Smoking history, onset before 50 years, infrapopliteal arterial
    occlusions, upper limb involvement, absence of atherosclerotic risk
    factors besides smoking

**What is important about diagnosing Buerger's**

-   Typically a diagnosis of exclusion

-   Must rule out proximal embolic source, trauma, local lesions (eg pop
    entrapment or cystic adventitial disease), autoimmune disease,
    hypercoagulable status, atherosclerosis

**What physical exam and non-invasive/invasive imaging findings of
Buerger's?**

-   Distal, but not proximal arterial disease (palpable
    brachial/popliteal but absent/reduced at ankle or wrist)

-   DBI\<0.6 and flat/reduced digital waveforms

-   CTA/MRA/DSA or duplex - characteristic findings of
    serpiginous/corkscrew collaterals, occlusion of distal calf/pedal
    arteries, and normal proximal arteries.[@busch2011; @fujii2011]

**What is the mainstay treatment in Buerger's disease?**

1.  Smoking cessation! Only treatment to improve symptoms and reduce
    amputation risk if achieved before onset of gangrene or tissue loss.
    Important to remember following treatments will likely fail without
    smoking cessation.

2.  If smoking cessation does not improve, medical management with
    antiplatelet agents, immunomodulators, vasodilators, anticoagulants

3.  Endovascular-distal small vessel intervention

4.  Surgical-upper extremity autogenous vein bypass-limited success due
    to poor outflow

5.  Sometimes can consider upper extremity sympathectomy, but unproven
    benefit

6.  Amputation-reported in 30-40% who are followed longer than 5
    years[@olin2018]

### Large Artery Vasculitis {#large-vasculitis}

**What are common characteristics for patients who are suspected to have
a large vessel vasculitis?** [@shanmugam137VasculitisOther2019;
@weyand2003]

-   Affect aorta and major branches

-   Present with non-specific heterogenous symptoms making the diagnosis
    challenging. Most commonly, they present with systemic or
    constitutional symptoms (fatigue, fever, weight loss, arthralgias)

-   Frequently, diagnosis made with presence of constitutional symptoms,
    elevated inflammatory markers, and dedicated imaging (MRA, CTA, DUS,
    or PET)

**How can you differentiate takayasu arteritis vs giant cell
arteritis?**

1.  Takayasu arteritis[@ehlert2019]

    1.  Aorta and primary

    2.  Young patients \<20 years and female in 80-90% of cases, Asian
        populations

    3.  Criteria (ACR)

        1.  Onset \<40 years

        2.  Claudication of an extremity

        3.  Decreased brachial pulse

        4.  \>10 mmHg SBP between arms

        5.  Bruit over subclavian arteries or aorta

        6.  Arteriographic evidence of narrowing/occlusion in
            aorta/primary branches/or large upper/lower extremity
            arteries

    4.  Initial therapy should include steroids and close observation.
        If persistent severe symptoms, then consider surgical repair.
        Mid aortic syndrome likely needs open repair from uninvolved
        aorta-usually thoracic aorta to bifurcation bypass with jump
        graft to involved visceral vessels.[@ehlert2019]

2.  Giant cell arteritis

    1.  Aorta and main branches, but predilection for carotid artery
        branches[@bongartz2006]

    2.  Diagnosis:

        1.  Age at disease onset \> 50 years

        2.  New headache

        3.  Temporal artery abnormality

        4.  Elevated ESR (\>50) - can be normal in up to a quarter of
            cases.

        5.  Abnormal artery biopsy (gold standard test)

    3.  Other symptoms include jaw pain with mastication or visual
        changes

    4.  Associated with Polymyalgia rheumatic, characterized by morning
        stiffness in shoulders/hips occurring in 40-50% of patients

    5.  Arteriography/MRA/CTA/PET may be used to assess large vessel
        involvement - classic finding is smooth, tapering stenosis.

**How should patients be monitored with active large artery
vasculitis?**

-   Lab data tracked at least monthly for 6 months with close follow-up
    to ensure appropriate response to medical treatment and enable
    physicians to assess for adverse effects of medical treatment

-   Repeat tests after remission reached and imaging choice to evaluate
    large vessels (DUS/CTA/MRA)

**What is the medical treatment for GCA and when do you consider
surgical treatment?**

-   Medical-steroid therapy. In as many as 50% of patients who have a
    large vessel vasculitis refractory to glucocorticoid therapy alone,
    patients will trial immunomodulators or cytotoxic truxs (ie
    methotrexate, azathioprine, mycophenolate, tocilizumab, or
    leflunomide)

-   Intervention-once remission, treatment of symptomatic arterial
    lesions or those at aneurysm size threshold should be considered and
    as many as 50-70% with large vessel vasculitis will require
    intervention.

    -   Endovascular-angioplasty/stent/stent graft for large vessel
        vasculitis have all been described, however higher restenosis in
        endovascular compared to open treatment

    -   Open Surgery (gold standard)-lesions are long, fibrotic and
        therefore less amenable to endovascular treatment. Bypass grafts
        from aorta-CCA are the most common (CEA should be avoid due to
        pathology involved)

        -   Upper extremity bypass with autogenous vein to the brachial
            artery

        -   Aortic aneurysms should be managed with open
            surgery[@janssen2008]

### Other Vasculitis syndromes

Polyarteritis Nodosum - Focal necrotizing lesions primarily affecting
medium-sized muscular arteries, peak incidence in 40s.  Multiple
saccular aneurysms secondary to inflammatory destruction of vessel
media.[@shanmugam2019; @weyand2003]

Behçet Disease - recurrent oral pathos ulcers, genital ulcers and
uveitis.[@shanmugam2019; @weyand2003]

## Aneurysmal Disease

**How are subclavian aneurysms caused and how can they present?**
[@baig84UpperExtremity2019]

Etiology/Pathology:

-   Degenerative - atherosclerotic or due to aberrant right subclavian
    with degenerative changes in proximal subclavian known as "Kommerell
    diverticulum"

-   Traumatic - blunt, penetrating, iatrogenic with attempted catheter
    placement

-   Radiation induced, sometimes seen after radiation for breast
    cancer.[@mohan2016]

-   Thoracic outlet obstruction - no need to investigate for aTOS if
    there is another plausible cause.

Presentation

-   Rare, comprises 1% of peripheral aneurysms.
-   Exam-pulsatile supraclavicular mass or bruit, absent/diminished
    pulses, signs of microembolization ("blue finger")
-   Most discovered incidentally, however referred chest, neck, shoulder
    pain, upper extremity ischemia due to thrombosis or distal embolism,
    brachial plexus compression, hoarseness from right recurrent
    laryngeal nerve compression
-   Dysphagia from esophageal compression in aberrant right subclavian
    artery
-   Rarely rupture

**What are diagnostic studies and treatment modalities for subclavian
aneurysms?**

-   CXR-mediastinal mass may suggest neoplasm

-   MRA/CTA important to delineate extent of aneurysm and proximity to
    ipsilateral vertebral artery

Treatment:

-   Open Repair-resection/endoaneurysmorrhaphy with end to end (small
    aneurysms) or interposition prosthetic graft[@vierhout2010]

    -   Proximal-median sternotomy with supraclavicular fossa extension
        for adequate proximal control for right side, however
        supraclavicular with left anterolateral thoracotomy for left
        subclavian aneurysm

    -   Mid-Distal-supraclavicular/infraclavicular generally adequate
        for control where again resection of the clavicle may be needed

-   Endovascular Repair-transbrachial/transfemoral approach with covered
    stent[@maskanakis2018]

    -   Must consider vertebral artery origin. Can cover vertebral
        artery if contralateral vertebral artery is patent and of
        adequate size, however posterior circulation stroke may occur
        when the contralateral vertebral artery is highly stenotic,
        hypoplastic or occluded.

-   Hybrid Repair-embolization/coils of proximal subclavian artery
    combined with subclavian transposition or carotid-subclavian bypass

-   For aberrant subclavian artery aneurysm, resection or exclusion of
    the aneurysmal artery with vascular reconstruction of the subclavian
    artery is recommended. Especially in the setting of dysphagia
    lusoria, subclavian artery reconstructed by interposition graft
    where proximal anastomosis is on ascending aorta. Alternatively,
    left posterolateral thoracotomy for proximal aneurysm resection and
    right supraclavicular incision for reconstruction of subclavian
    artery by end to side to the right CCA has been reported.

**How are axillary aneurysms caused and how can they present?**

Etiology/Pathology:

-   Blunt/penetrating trauma

-   Congenital (infrequently reported)

-   Post-traumatic axillary aneurysms (repeated abduction/external
    rotation downward toward humeral head in baseball pitchers)

Presentation:

-   Exam-pulsatile supraclavicular mass or bruit, absent/diminished
    pulses, signs of microembolization ("blue finger")

**What are diagnostic studies and treatment modalities for axillary
aneurysms?**

Diagnosis:

-   Ultrasound

-   CTA/MRA of upper extremity

Treatment:

-   Open Repair-resection with interposition vein grafting or prosthetic
    if inadequate vein is present.

-   Endovascular repair-covered stent graft can be placed with
    occasional embolization with micro coils to isolate sac and prevent
    retrograde endoleaks

**How are brachial aneurysms caused and how can they present?**

Etiology/Pathology:

-   False aneurysms secondary to repetitive trauma

-   Iatrogenic complications

-   IV drug abuse (infected pseudoaneurysms in antecubital fossa)

-   Connective tissue disorders (ex. type IV Ehlers danlos)

Presentation:

-   Exam: pulsatile mass

-   Local pain or symptoms of median nerve compressions

-   Hand/digital ischemia from thrombosis/distal embolization

**What are diagnostic studies and treatment modalities for brachial
aneurysms?**

Seen in 1-3% of brachial artery access - increased risk in older age,
female patients, and larger sheaths.[@treitl2015] More on appropriate
access technique can be found in \@ref(endovascular)

Diagnosis:

-   Duplex Ultrasound

-   CTA/MRA of upper extremity may be needed to delineate extent of
    aneurysm

Treatment:

-   Open Repair (preferred)-resection with patch or interposition vein
    grafting

-   Endovascular repair-rare and generally in a traumatic setting

-   Iatrogenic injuries-due to access and nonoperative treatment for
    small/asymptomatic pseudoaneurysms that are likely to thrombose
    spontaneously. Direct suture repair with evacuation of hematoma is
    possible. Thrombin injection is less favorable due to location and
    short neck.

-   Neurologic symptoms often from median nerve compression an require
    urgent repair, open surgery best for decompression.

**How do you manage a patient who presents with a radial artery
pseudoaneurysm after a coronary angiogram on year ago?**

Small pseudoaneurysms \<3cm have a high rate of sponatenous thrombosis.
However, larger psuedoaneurysms or those with symptoms require
treatment. Often best managed with open excision and primary repair or
interposition graft.[@tosti2017]

Can also be seen as a result of trauma.[@bagir2017]

## Occupational Vascular Disease

**There are some occupational vascular disorders than contribute to
vascular disease in the upper extremity. Hand arm vibration syndrome and
hypothenar hammer are of particular importance. Can you talk to us about
the key information from these syndromes?**
[@eskandari185ConditionsArising2020]

### Hand-Arm Vibration Syndrome

Etiology:

-   Vibrating handheld machines (eg pneumatic hammers and drills,
    grinders, and chain saws)

-   Linear relationship between exposure over years and onset of this
    syndrome

-   Exact mechanism unknown, but thought that endothelial damage with
    sympathetic hyperactivity -\> finger blanching attack

Presentation:

-   Various stages seen where early results in slight tingling/numbness
    and lateral, the tips of one or more fingers experience attacks of
    blanching that is usually precipitated by cold

-   Blanching typically lasts 1 hour and terminates with reactive
    hyperemia, but prolonged exposure can cause bluish black cyanosis of
    fingers

Diagnosis

-   Detailed history with use of vibrating tools/symptoms of Raynaud
    phenomenon

-   Objectively: cold induced ischemia with recording time until digital
    temperature recovers

-   Digital occlusion with noninvasive digit pressures or duplex
    scanning

Treatment

-   Avoidance of vibratory tools

-   Nifedipine (Ca2+ channel blocker) in advanced cases

-   IV prostanoid (ie prostacyclin) for digital gangrene

-   Surgery-cervical sympathetectomy or digital sympathectomy rarely
    needed\

### Hypothenar hammer syndrome

Etiology:

-   Repetitive use of palm of hand in activities that involve pushing,
    pounding, twisting - particularly to the ulnar artery as it exits
    guyots canal and crosses the hook of the hamate.[@ferris2000]

-   Name comes from reports of mechanics, factory workers, carpenters or
    laborers who habitually use there hands as a hammer are ad risk for
    disease

-   Repetitive trauma leads to thrombotic occlusion, aneurysm formation
    or both

Presentation:

-   Asymmetrical distribution involving dominant upper extremity where
    cyanosis and pallor can occur and digits affected are ulnar
    distribution in nature

-   Cool/mottled digits or severe cases with ischemic ulcers

Diagnosis:

-   Duplex ultrasound

-   CTA or MRA

-   Arteriography (gold standard) with corkscrew pattern typically in
    affected vessels

Treatment

-   Conservative-smoking cessation/hand protection/cold
    avoidance.[@carr2019]

-   Medical-calcium channel blockers/antiplatelet

-   Surgical (severe digital ischemia/aneurysm)-ligation if adequate
    collateral or interposition vein graft has good long term patency.

### Environmental Exposures

**Exposure to what environmental agents can result in upper extremity
ischemia?**

Acrosteolysis

-   Exposure to polyvinyl chloride can result in ischemic hand symptoms
    similar to those of Raynaud syndrome

-   Angiography-damage to digital arteries with multiple
    stenosis/occlusions or hyper vascularity adjacent to areas of bone
    resorption

-   Treatment-supportive

Electrical burns

-   \<1000 V cause injuries limited to immediate skin/soft tissue,
    however \>1000 V cause damage from entry to exit point

-   Results in arterial necrosis with thrombus or bleeding and gangrene
    of digits develop

-   Initially can be occlusion/thrombosis or spasm, however later damage
    can cause aneurysmal degeneration

-   Treatment-dependent on soft tissue/bone injuries as well. Can have
    reconstruction with free flap due to local vascular damage or
    occlusion of major artery requiring bypass grafting

Extreme thermal injuries

-   Workers at risk with chronic exposure to cold (slaughterhouse,
    canning factory, and fisheries)

-   Raynaud syndrome symptoms due to vasomotor disturbances in the hands
    when exposed to extreme chronic thermal trauma

-   Treatment-Supportive

### Sports Medicine

**How can athletes specifically be affected by upper extremity
ischemia?**

Overview

-   Athletes who engage in strenuous or exaggerated hand/shoulder
    activity may be susceptible to upper extremity ischemia from
    arterial injury manifested by Raynaud syndrome, symptoms of sudden
    arterial occlusion or digital embolization

## Vascular Trauma-Upper Extremity

**This is discussed in detail here:** \@ref(peripheral), **so we will go
over some important specifics for upper extremity vascular injury.**
[@kauvar184VascularTrauma2020]

**What is the management strategy for an iatrogenic placement of a
subclavian artery catheter?**

Critically ill patients can be treated effectively with the use of a
closure device or covered stent placement.[@yoon2015; @tran2009;
@cohen2014] Cut down in the mid subclavian artery is very difficult to
perform and results in increased blood loss and worse outcomes.

**What is the mechanism and management of upper extremity axillary
artery trauma?**

Mechanism and Pattern

-   Predominantly in penetrating trauma with equal incidence in
    proximal/middle/distal divisions and brachial plexus injury in
    \>1/3rd of arterial injury

Diagnostic Considerations

-   Physical exam with deficiencies in upper extremity pulses/ischemic
    changes, but may not be present given collateral flow from axillary
    artery to upper extremity

-   High index of suspicion with location of injury proximity to course
    of axillary artery

-   Upper extremity Doppler or CTA if patient is stable for diagnosis

Surgical Considerations

-   Primary repair or treated with interposition graft

-   If hemodynamically stable, can consider covered stent based on
    location to thoracic outlet via femoral/brachial approach

**What is the mechanism and management of upper extremity brachial
artery trauma?**

Mechanism and Pattern

-   Frequently associated with humerus fractures/elbow dislocation

-   Penetrating trauma

Diagnostic Considerations

-   Pulse deficit in majority (\>75% of cases)

-   Upper extremity Doppler of CTA

Surgical Considerations

-   Given course, can be extensively mobilized and repaired in
    end-to-end fashion in 50% of cases. Otherwise, treatment with an
    interposition graft

**What is the mechanism and management of upper extremity radial/ulnar
artery trauma?**

Mechanism and Pattern

-   Associated with significant soft tissue pattern

Diagnostic Considerations

-   Pulse deficit in \>80% of patients

-   Doppler based Allen test-confirm radial/ulnar contribution to palmar
    arch

Surgical Considerations

-   If Allen test reveals a patent palmar arch, the injured artery can
    be ligated

-   If palmar arch is not patent in the absence of contribution of the
    injured artery, it should be repaired

-   If both are damaged, preference to ulnar artery as dominant
    contribution to hand

-   Generally, repair can be done in an end to end fashion given
    mobility of the vessel

## Compression Syndromes

**The main syndromes are quadrilateral space syndrome and humeral
compression of the axillary artery. What important information here do
our listeners need to know?**

### Quadrilateral space syndrome

Anatomy:

-   Bordered by teres minor superiorly, humeral shaft laterally, and
    teres major inferiorly, and long head of triceps muscle medially

-   Posterior humeral circumflex artery and axillary nerve in space

Pathophysiology

-   Compression of posterior humeral circumflex occurs with
    abduction/external rotation

-   Typically seen with chronic overhand motion athletes
    (pitchers/volleyball players)

-   Vascular-repetitive mechanical trauma to posterior circumflex
    humeral artery

-   Neurogenic-fixed structural impaction of quadrilateral space by
    fibrous bands or space-occupying lesions

Presentation

-   Muscle atrophy, paresthesias, poorly localized shoulder pain and
    pain in quadrilateral space

Treatment

-   Medical: Oral anti-inflammatory medications, PT, limitation of
    activities

-   Surgery: decompression with neurolysis/excision of fibrous bands or
    other space occupying lesions

### Humeral head compression of axillary artery

Anatomy:

-   3rd portion of axillary artery compressed by head of humerus

Etiology/Pathophysiology:

-   Arm is abducted and externally rotated with downward compression of
    humeral head to axillary artery

Presentation:

-   Arm fatigue, loss of pitch velocity, finger numbness, Raynaud,
    cutaneous embolization

Diagnosis:

-   Provocative maneuvers with impedance of flow through axillary artery
    on ultrasonography

-   Arteriography with rest and provocative position

Treatment:

-   Supportive with avoidance of throwing motion

-   Surgical-saphenous vein patch for no improvement or structural
    injury may require resection with saphenous vein bypass anatomically
    or extra-anatomic tunneling above pec minor

## Thoracic Outlet Syndrome

**27 Nov 2019:** *Nedal Katib, Prince of Wales, Sydney Australia*

```{r echo=FALSE}
library(embedr)

embed_audio("https://traffic.libsyn.com/audiblebleeding/VSITE_TOS_02auph.mp3")

```

Thoracic Outlet Syndrome = A constellation of signs and symptoms
relating to the compression of the neurovascular structures that occurs
as these structures travel between the thoracic aperture and the upper
limb.

Types: Neurogenic, Venous and Arterial

-   vTOS -- 2-3%

-   aTOS -- 1%

-   nTOS -- \>95% [@humphries124ThoracicOutlet2019]

### Anatomy

Understanding the anatomy of what is collectively referred to as the
thoracic outlet is the best way to thoroughly appreciate this topic.

Anatomy from anterior to posterior

-   Subclavian vein

-   Phrenic nerve

-   Anterior scalene muscle attachment to the first rib

-   The subclavian artery

-   The brachial plexus

-   The middle scalene muscle.

Three spaces where the neurovascular structures are at risk of
compression:

1.  Interscalene Triangle

2.  Costoclavicular Passage[@garygwindAnatomicExposuresVascular2013]

3.  Subcoracoid Space[@garygwindAnatomicExposuresVascular2013]

**Interscalene Triangle:**

Appreciating the attachments of the Anterior and Middle Scalene Muscles
on the first rib becomes important in the diagnosis of the various types
and also the ultimate surgical management of the compression.

**Anterior Scalene:**

Attachments: Anterior Tubercles of the four 'typical' cervical vertebrae
(3-6) AND the scalene tubercle on the upper surface of the first rib.

-   Phrenic nerve runs along anterior scalene muscle and injury can
    cause ipsilateral diaphragm paralysis.

**Middle Scalene:**

Attachments: The posterior tubercles and intertubercular lamellae of all
the cervical vertebrae AND the Quadrangular area between the neck and
subclavian groove of the first rib. [@mcminnLastAnatomyRegional2019]

-   Long thoracic nerve runs along middle scalene muscle and injury can
    cause winged scapula.

**The First Rib:**

-   The broadest and flattest of the ribs and is an 'Atypical Rib'.

-   The upper surface of the first rib has the scalene and quadrangular
    tubercles for attachments of the anterior and middle scalene muscles
    respectively. There are also three grooves for the Subclavian Vein,
    artery and the Lower Trunk of the Brachial Plexus.

-   The Inferior Surface is smooth and inferior and medially has an
    attachment for the suprapleural membrane, Sibson's fascia AKA
    scalenus minimus, which is tethered to the C7 vertebrae.

-   This is the passage of the subclavian vein largely as it emerges
    through the tight space created by the clavicle, the subclavius
    muscle and the costoclavicular ligament and also more posteriorly
    this can also compress the artery and nerves as the space can also
    be narrow in relation with the scapula and subscapularis.
    [@garygwindAnatomicExposuresVascular2013]

**Subclavius Muscle:**

-   Attached to the costochondral junction of the first rib and is
    inserted into the subclavian groove on the inferior surface of the
    clavicle. [@mcminnLastAnatomyRegional2019]

-   This space is best appreciated by intimate knowledge of three
    things:

    -   The Coracoid Process and its attachments

    -   The Pectoralis Minor Muscle

    -   The Clavipectoral Fascia

**The Coracoid Process:**

-   Arising from the Scapula as a 'process', this broad-based bony
    landmark offers attachment to muscles and ligaments.

-   The relevant attachments being the pectoralis minor muscle occupying
    the medial border for about 2cm behind its tip. The tip itself
    having a medial and lateral facet for the short head of biceps and
    the coracobrachialis muscles respectively.

**Pectoralis Minor Muscle:**

Attached to the bone of the third, fourth and fifth ribs AND the medial
border of the coracoid process.

**Clavipectoral Fascia:**

A sheet of fascial membrane that fills the space between the clavicle
and pectoralis minor splitting and encompassing the subclavius muscle.
Its superior portion is what can be thickened and become a tight band
referred to as the costocoracoid ligament.

**Phrenic Nerve Anomaly:**

The Phrenic Nerve normally runs anterior to the Subclavian Vein. A rare
anomaly is the nerve compressing the vein anteriorly and in very rare
circumstances due to the timing of development can run through the vein
itself.

Anomalous anatomy can also cause TOS especially when patients have a
Cervical Rib and anomalous first ribs or a congenital band attaching to
the first rib.

-   Incidence of anomalous first ribs and cervical ribs is 0.76% and
    0.75% respectively.

-   Incidence of bands are as high as 63% in the general population.
    [@humphries124ThoracicOutlet2019]

nTOS

-   Scalene Triangle compression -- most common cause of brachial plexus
    and neurogenic TOS

-   Cervical Rib and Anomalous First Rib

aTOS

-   Cervical Rib and Anomalous First Rib

-   Scalene Triangle compression

vTOS

-   Costoclavicular Passage

-   Subcoracoid Space

### Diagnosis and Evaluation

#### Patient History

-   Identify symptoms and thoroughly interrogate timing

-   Exclude history of trauma

-   Associated symptoms like headache, visual disturbance, neurology in
    the upper limb

-   Exclude Carpal Tunnel and Antecubital Tunnel Syndromes if symptoms
    are isolated to the arm or forearm or hand

-   Patients with vTOS may present acutely and have acute or subacute
    Upper Limb DVT

-   Patients with aTOS need to be investigated and assessed urgently
    given risk of ischemia.

#### Clinical Examination

Provocative maneuvers are largely used for nTOS. While these are
described and mentioned in most texts their utility largely is beyond
the scope of a vascular surgeon's assessment and diagnosis of nTOS.

**Adson Test**

-   Extended abducted and externally rotated arm -- palpate radial pulse

-   Rotate and laterally flex the neck to the ipsilateral side while
    inhaling deeply.

-   A positive test results in reduction or complete obliteration of
    radial pulse

**Roos Test / EAST test**

-   Patient seated and both arms abducted 90 degrees and externally
    rotated and elbows flexed at 90 degrees.

-   Open and close hands for 3 minutes or until pain or paraesthesia
    sets in.

**Elveys Test**

-   Abduct both arms to 90 degrees with elbows extended and dorsiflex
    both wrists.

-   If pain is elicited as wrists dorsiflexed then test is positive.

-   A further manoeuvre is then performed, laterally flex the head on
    each side, if pain is elicited on the contralateral side to which
    the head is flexed then test is positive.
    [@humphries124ThoracicOutlet2019]

#### Non-invasive imaging or vascular lab studies

-   DBI

-   Arterial Duplex

-   Venous Duplex

-   CT -- CTV commonly performed in acute upper limb DVT and suspicion
    of vTOS

-   CTA for the evaluation of aTOS and excluding other causes of
    embolisation

-   MRI -- for further evaluation of the anatomy and related
    neurovascular compression

-   Electromyography and Nerve Conduction Studies for nTOS

**Paget Schroetter Syndrome**

-   First defined by Hughes in 1949 in reference to Sir James Paget who
    in a hundred years earlier defined acute arm swelling and pain as
    possibly related to vasospasm and then von Schroetter who in 1884
    attributed to the presentation to subclavian and axillary vein
    thrombosis. [@humphries123ThoracicOutlet2019]

-   Now vTOS and Paget Schroetter Syndrome are used synonymously.

-   Paget Schroetter Syndrome accounts for 10-20% of all upper extremity
    deep vein thrombosis. [@sekharYearbookVascularEndovascular2018]

### Rib Resection approaches

+--------------------+--------------------+--------------------+
|                    | Advantages         | Disadvantages      |
+====================+====================+====================+
| Tran saxillary     | Cosmetically more  | -   Difficult to   |
|                    | appealing as it    |     visualize the  |
|                    | has a limited      |     anatomy,       |
|                    | hidden scar        |     dependent on   |
|                    |                    |     good           |
|                    |                    |     assistance     |
|                    |                    |                    |
|                    |                    | -   Risk of injury |
|                    |                    |     to T1 nerve    |
|                    |                    |     root, phrenic  |
|                    |                    |     nerve, long    |
|                    |                    |     thoracic,      |
|                    |                    |     brachial       |
|                    |                    |     plexus ,       |
|                    |                    |     subclavian     |
|                    |                    |     vein and       |
|                    |                    |     arterial with  |
|                    |                    |     limited        |
|                    |                    |     exposure to    |
|                    |                    |     repair         |
|                    |                    |                    |
|                    |                    | -   Not able to    |
|                    |                    |     approach       |
|                    |                    |     cervical ribs, |
|                    |                    |     scalene        |
|                    |                    |     triangle or    |
|                    |                    |     patch vein.    |
+--------------------+--------------------+--------------------+
| Suprac lavicular   | -   Good for       | -   Unable to      |
|                    |     scalene        |     decompress     |
|                    |     triangle       |     venous         |
|                    |     access and     |     compression or |
|                    |     debulking and  |     visualize vein |
|                    |     cervical rib   |     adequately     |
|                    |     resection      |                    |
|                    |                    | -   Cosmetically   |
|                    | -   Required for   |     less appealing |
|                    |     aTOS if        |                    |
|                    |     arterial       |                    |
|                    |                    |                    |
|                    | reconstruction     |                    |
|                    | necessary          |                    |
+--------------------+--------------------+--------------------+
| Infrac lavicular   | -   Good access    | -   Unable to      |
| [\@ siracuseI      |     for venous     |     expose         |
| nfraclavi          |                    |     subclavian     |
| cularFirs          | decompression      |     artery or      |
| tRib2015]          |                    |     decompress     |
|                    | -   Allows for     |     brachial       |
|                    |     excision of    |     plexus.        |
|                    |     subclavius     |                    |
|                    |     muscle and     | -   Difficult to   |
|                    |                    |     access most    |
|                    | c ostoclavicular   |     posterior      |
|                    | ligament           |     aspect of rib  |
|                    |                    |                    |
|                    |                    | -   Cosmetically   |
|                    |                    |     less appealing |
+--------------------+--------------------+--------------------+
| Parac lavicular    | -   Useful if      | -   Requires two   |
|                    |     mixed etiology |     incisions one  |
|                    |     TOS to         |     above and      |
|                    |     adequately     |     below the      |
|                    |                    |     clavicle       |
|                    | decompression all  |                    |
|                    |                    |                    |
|                    | neurovascular      |                    |
|                    | structures         |                    |
+--------------------+--------------------+--------------------+

#### Post operative complications

-   Post operative patients with hemodynamic instability and ipsilateral
    effusion on xray should go back to OR for exploration and hemorrhage
    control. @rinehardtCurrentPracticeThoracic2017

-   Chyle leak often managed with adequate drainage and medium chain
    fatty acid diet. Persistent leak should undergo wound exploration
    and ligation of thoracic duct (most often encountered on the left)
    or VATS thoracic duct ligation.[@delaney2017] IR embolization has
    also been shown to be effective.[@cope1999]

### vTOS

#### Demographics

-   Incidence: 2/100,000 persons

-   Age: 18 years to 30 years
    [@illigComprehensiveReviewPagetSchroetter2010]

-   M\>F

#### Presentation

-   Upper Limb edema, pain and cyanosis. Edema affects the shoulder, arm
    and hand and characteristically non pitting.

-   Collateral vein dilatation over the shoulder, neck and anterior
    chest wall to accommodate for the increased venous hypertension.
    [@humphries123ThoracicOutlet2019]

-   Pain on exertion of the upper limb described as stabbing, aching or
    tightness.

-   The reported incidence of PE following Upper Limb DVT is \<12%.
    [@humphries123ThoracicOutlet2019]

#### History

-   A differential diagnosis for Upper Limb DVT

    -   vTOS

    -   Congenital Phrenic Nerve anomaly

    -   History of Fracture, Clavicular Fracture and malunion

    -   Repetitive arm provocative manoeuvres, check occupation and
        history of body-building

        -   Pectoralis Minor Hypertrophy.

-   Exclude Pulmonary Embolism

-   Exclude Venous Gangrene and Phlegmasia of the upper limb

#### Goals of therapy for vTOS

Limited evidence due to lack of RCT's. Majority of evidence based on
retrospective studies.

-   Prevent immediate risk

-   Return patient to unrestricted use of the affected extremity

-   Prevent recurrence of thrombosis without the need of long-term
    anticoagulation

-   Prevent long term Post Phlebitic Limb Syndrome

#### Initial management strategy

-   As per ACCP Guidelines: Initial management is anticoagulation
    regardless of etiology. [@kearonAntithromboticTherapyVTE2016]

    -   The limitations of anticoagulation alone are that the slow
        recanalization of the thrombus may lead to eventual valvular
        damage and intravenous scarring.
        [@sekharYearbookVascularEndovascular2018]

    -   Thrombolysis has been considered superior to anticoagulation
        alone in minimizing valvular damage due to residual clot.
        [@urschelSurgeryRemainsMost2008]

    -   Systemic Lysis -- non favored due to risk of intracranial
        hemorrhage. [@grunwaldCatheterDirectedThrombolysisTreatment2004]

    -   Catheter Directed Lysis (CDT) -- carries a lower risk of
        intracranial hemorrhage.

    -   Patient should be maintained in a compression sleeve until
        definitive decompression can be performed.

-   Optimal timing of CDT

    -   Within 14 days of onset of thrombosis. Excellent results have
        been reported following CDT if initiated before 14 days.
        [@wilsonFibrinolyticTherapyIdiopathic1990]

-   Surgical indications for vTOS

    -   After initial management patients are generally divided into two
        groups, unsuccessful or successful thrombolysis.

    -   Persistent stenosis or signs of extrinsic compression, on
        venography, has generally been perceived as a significant risk
        of recurrent thrombosis.

    -   Surgery for vTOS remains to be mainly Rib Resection and
        decompression of the subclavian vein with or without venolysis
        and patch plasty either surgical or endovenous.

    -   Surgical treatment of severe resistant subclavian vein stenosis
        in the setting of vTOS is rib resection by paraclavicular
        approach and vein patch plasty.
        [@melbyComprehensiveSurgicalManagement2008]

    -   Venous occlusion in vTOS may be treated with jugular turn down
        or venous bypass to IJ of SVC if patients remain symptomatic.
        [@vemuriDiagnosisTreatmentEffortinduced2016]

#### Controversy around vTOS

-   There is a lack of consensus around the necessity of surgical rib
    resection, the timing and the requirement for vein patch plasty.

-   Options post recanalization:

    -   Deferring surgical decompression for 1-3 months after
        thrombolysis to allow for healing of the venous endothelium and
        resolution of the acute inflammatory process.
        [@humphries123ThoracicOutlet2019]

    -   Decompression during the same admission, as the thrombolysis,
        with the main benefit being to reduce the risk of re-occlusion.
        [@humphries123ThoracicOutlet2019;
        @molinaPagetSchroetterSyndromeTreated2007]

    -   Post decompression venography and treatment 2 weeks post rib
        resection may help to prevent recurrence and long term vein
        patency. [@changRoutineVenographyFollowing2012]

#### Landmark papers regarding vTOS and what are the take home messages

1.  Lugo J et al -- Acute Paget Schroetter syndrome: does the first rib
    routinely need to be removed after thrombolysis? Annals of Vascular
    Surgery 2015 [@lugoAcutePagetSchroetter2015]

    1.  Systematic literature review analysis. Patients divided into
        three groups

        1.  First Rib resection (FRR) -- n=448

        2.  First Rib resection and endovenous venoplasty (FRR and
            PLASTY) n=68

        3.  No further intervention after Thrombolysis n=168

    2.  Symptom relief after initial follow up more likely in FRR (95%)
        and FRR and PLASTY (93%) compared to no rib removed (54%) --
        p\<0.0001

    3.  Results showed superior patency with FRR and PLASTY and FRR
        compared to anticoagulation alone.

    4.  Conclusion was that patients are more likely to experience
        greater long-term results with FRR compared to no FRR.

2.  Sajid MS et al -- Upper limb vein thrombosis: a literature review to
    streamline the protocol for management. Acta Haematology 2007
    [@sajidUpperLimbDeep2007]

    1.  a comprehensive review identifying the key papers on this topic
        and allows for a clear view of the best management strategy.

3.  Vemuri, C., Salehi, P., Benarroch-Gampel, J., McLaughlin, L. N., &
    Thompson, R. W. (2016). Diagnosis and treatment of effort-induced
    thrombosis of the axillary subclavian vein due to venous thoracic
    outlet syndrome. *Journal of Vascular Surgery: Venous and Lymphatic
    Disorders*, *4*(4), 485--500.
    [@vemuriDiagnosisTreatmentEffortinduced2016]

    1.  Comprehensive summary of management strategy for effort induced
        thrombosis.

### aTOS

#### Presentation

-   Most common: Hand ischemia due to arterial compression or
    microembolization with subclavian artery aneurysm and pulsatile
    supraclavicular mass [@boll122ThoracicOutlet2019]

-   Less common: Exertional pain, unilateral Raynaud's Phenomena,
    retrograde embolisation and neurological symptoms

-   Clinical Examination

    -   Audible Bruit / Palpable thrill over the supraclavicular fossa

    -   Pulsatile mass

    -   Distal ischemic lesions in the distal hand -- Splinter
        hemorrhages

    -   Positive Adson Test

-   Differential Diagnosis

    -   Trauma

    -   Primary and Secondary Raynaud's Phenomena

    -   Small Vessel Vasculitis

    -   Connective Tissue Disorders

    -   Thromboangiitis Obliterans

    -   Arterial Embolisation -- Aortic or Central Source

    -   Radiation Arteritis

    -   Atherosclerotic / Dissection causes

-   The different anatomical abnormalities causing aTOS
    [@boll122ThoracicOutlet2019]

    -   Cervical Rib (60%)

    -   Anomalous First Rib (18%)

    -   Fibrocartilaginous band (15%)

    -   Clavicular Fracture (6%)

    -   Enlarged C7 transverse process (1%)

#### Scher Staging of aTOS

-   Stage 0: Asymptomatic

-   Stage 1: Stenosis of Subclavian Artery with minor post stenotic
    dilatation with no intimal disruption

-   Stage 2: Subclavian artery aneurysm with intimal damage and mural
    thrombus

-   Stage 3: Distal embolisation from subclavian artery disease

#### Diagnosis

Most useful studies are pulse volume recordings (PVR) and duplex to
identify aneurysm or sites of embolization. Stress test is not reliable
for diagnosis. [@vemuriClinicalPresentationManagement2017;
@criadoSpectrumArterialCompression2010]

#### Management considerations with aTOS

-   Symptomatic patients are generally indicated for treatment. Unlike
    asymptomatic patients in whom it may be appropriate to manage
    conservatively. [@boll122ThoracicOutlet2019]

-   Supraclavicular rib resection is the most suitable for adequate
    arterial reconstruction. Transaxillary has been argued to offer more
    complete rib resection however arterial repair is not possible in
    this approach.

-   Subclavian artery repair is necessary in Scher Stages 2 and 3 and in
    some cases 1. Arterial repair with conduit either GSV, Femoral Vein
    or prosthetic have been described. Ringed PTFE offers good patency
    and resistance to kinking in this functional anatomical location.

### nTOS

#### Demographics

Neurogenic TOS is largely a clinical diagnosis with symptoms and signs
pertaining to nerve compression most commonly the lower trunk of the
brachial plexus.

-   F\>M -- 70% Female

-   Ages 20-40

-   Occupational Exposure

-   Trauma history - often associated with recent history of neck
    trauma.

#### Presentation of nTOS

-   Symptoms [@sadeghi-azandaryaniThoracicOutletSyndrome2009;
    @sandersDiagnosisThoracicOutlet2007]

    -   Paraesthesia (98%)

    -   Trapezius pain (92%)

    -   Neck, shoulder or arm pain (88%) - usually involves nerve roots
        so isolated hand symptoms is rare.

    -   Supraclavicular pain with or without occipital headache (76%)

    -   Chest pain (72%)

    -   Weakness

    -   Swelling

-   Positional Effects

    -   Reproducible exacerbation of symptoms

    -   Lying supine with arms overhead

    -   Overhead activities -occupational or recreational

-   Weakness and Muscle Atrophy

    -   Hypothenar atrophy

    -   Drop-off in athletic performance

    -   Inability to carry out activities of daily living

#### The role of the Vascular Surgeon with nTOS

Often these patients have already seen multiple specialists and
physiotherapists.

-   Exclude other causes

-   Confirm diagnosis -- Neurophysiologic Tests (EMG and NCS)

-   Seek alternate opinion

-   Trial of Physiotherapy and non-operative management - patients
    should be evaluated and undergo a 6 week course of physical therapy.
    This physical therapy focuses on scalene and pectoralis stretching
    improving mobility of the shoulder and strengthening the arm. Many
    improve with physical therapy.
    [@baldermanPhysicalTherapyManagement2019]

-   Anterior scalene lidocaine block may provide temporary symptom
    relief (\~7 days) and may help identify those patients most likely
    to benefit from surgical decompression.
    [@salhanPC214UltrasoundGuidedAnesthetic2016;
    @lumImpactAnteriorScalene2012]

-   Botulinum injection may give an average of 6 weeks of relief.
    [@salhanPC214UltrasoundGuidedAnesthetic2016]

-   Be selective in patients who may require surgery

Surgery with Rib resection often is accomplished with transaxillary or
supraclavicular approach, particularly if scalenectomy or cervical rib
resection is necessary.

## Lymphedema

#### Etiology

Often due to breast cancer therapy and axillary lymph node
dissection.[@morrell2005]

### Management

Initial treatment includes decongestive therapy, such as daily massage
for 2-4 weeks. Once the girth and symptoms stabilize, then can
transition to long term management with compression sleeve.[@mondry2004]

Severe cases my require lymph node transfer.[@warren2007]

Lymphangiosarcoma in the setting of lymphedema (Stewart-Treves syndrome)
without distant disease is best managed with wide local excision and
adjuvant chemoradiation. Prognosis is poor.[@sharma2012]
