---
editor_options: 
  markdown: 
    wrap: 72
---

# Lower Extremity Occlusive Disease

## Pathophysiology

24 Aug 2020: Nedal Katib and Danielle Bajakian

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/PAOD_part_1_postEQ.mp3")

```

**What is Peripheral Arterial Disease (PAD) and what does it
encompass?**

Peripheral Arterial Disease encompasses extremity arterial disease but
generally is used to describe lower limb arterial occlusive disease.
PAOD is a more specific term and this encompasses atherosclerotic
disease of the lower limb arteries.

The disease has various presentations on a spectrum of asymptomatic
disease to intermittent claudication and finally chronic limb
threatening ischemia (CLTI) formerly known as Critical Limb Ischemia
(CLI) [@aboyans2017ESCGuidelines2018]

**What is the underlying pathophysiology of PAOD?**

The underlying pathophysiology which results in occlusive arterial
disease of the lower limb has somewhat evolved over the last 50 years.
While atherosclerosis remains, the main pathological process resulting
in occlusive disease, smoking related atherosclerosis has in the last 30
years been confounded with the rising incidence of diabetes and in
addition an aging population with progressive arterial disease.

Atherosclerosis, in summary, begins with an injury to the intimal lining
of the arterial wall, which can result from smoking, hypertension or
advanced age and ultimately a chronic inflammatory reaction resulting in
plaque build up and calcification that may result in progressive
stenosis and occlusion or plaque rupture with acute occlusion.

**What are the risk factors for atherosclerosis?**

Modifiable:

1.  ***Smoking***

    -   The **most significant modifiable risk factor** for developing
        peripheral arterial disease

    -   Causes Endothelial dysfunction by reducing nitric oxide and
        triggering reactive-oxygen species
        [@unitedstatessurgeongeneralHealthConsequencesSmoking2014]

    -   Causes a prothrombotic environment by causing an increase in
        thromboxane A2 and decreasing Prostacyclin thus overall
        resulting in an increased prothrombotic environment for
        platelets.

    -   Smoking has a stronger association with Intermittent
        claudication than with Coronary Artery Disease!
        [@gordonPredispositionAtherosclerosisHead1972]

2.  ***Diabetes , Metabolic Syndrome and Insulin Resistance***

    -   Diabetes Mellitus, after smoking, is the most significant
        modifiable risk factor for developing peripheral arterial
        disease. Both insulin resistance and hyperinsulinemia are
        independent risk factors for developing peripheral arterial
        disease.

    -   The Odds Risk for developing PAD in patients with DM ranges from
        1.89 to 4.05.

    -   An increase in HbA1C by 1% correlates with a 28% increase risk
        of developing PAD. [@adlerUKPDS59Hyperglycemia2002]

3.  ***Hypertension***

    -   The **most common** cardiovascular risk factor worldwide.

    -   The Incidence of PAD increases to 2.5-fold in patients with
        Hypertension. [@kannelUpdateEpidemiologicFeatures1985]

4.  ***Dyslipidemia***

    -   A strong association has long been identified as a risk factor
        for cardiovascular disease.

    -   25% cardiovascular event reduction for each 39 mg/dL (1mmol/L)
        reduction in LDL.
        [@heartprotectionstudycollaborativegroupRandomizedTrialEffects2007]

Non Modifiable:

1.  **Age**

    -   Age is identified as a risk factor for PAD regardless of gender.

    -   Prevalence of PAD increases with age: 15% \> 70 years of age.

2.  **Gender**

    -   The Framingham Study has found that the risk of developing PAD
        is doubled in men.

3.  **Ethnicity**

    -   The MESA study showed a higher prevalence of PAD (ABPI \<0.9) in
        African Americans compared to Whites. 7.2% versus 3.6%.
        [@bildMultiEthnicStudyAtherosclerosis2002]

    -   Cross Sectional analysis 6653 subjects all with ABPI assessment
        revealed a prevalence of PAD (\<0.9) of 4%. Non-Hispanic Whites:
        3.6%, Asian: 2%, African American: 7.2% and Hispanic: 2.4%.
        (p\<0.01) [@allisonEffectNovelCardiovascular2006]

**What are the some of the major population-based trials looking at the
natural history?**

1.  The Framingham Heart Study: The original Cohort from the town of
    Framingham, n=5183 patients followed over time for over 30 years.
    There have been multiple subsequent recruited populations since. The
    majority of information we have about risk factors related to
    cardiovascular health comes from this study.
    [@mahmoodFraminghamHeartStudy2014]

2.  The Rotterdam Study: 1990, Longitudinal Study, \>7000 particpants.

3.  CVHS -- 1989-1999 Longitudinal Study : n\>5000 Multicentre Study.

4.  MESA -- Cross Sectional analysis 6653 subjects all with ABPI
    assessment revealed a prevalence of PAD (\<0.9) of 4%. Non-Hispanic
    Whites: 3.6%, Asian: 2%, African American: 7.2% and Hispanic: 2.4%.
    (p\<0.01) [@allisonEffectNovelCardiovascular2006]

5.  The Edinburgh Study: The EAS began as a cross sectional study of
    1592 men and women in Edinburgh with the goal of examining the
    frequency of risk factors for peripheral arterial disease. The
    subjects were followed over 20 years.
    [@fowkesEdinburghArteryStudy1991]

**How does Diabetes confound the clinical picture of PAOD?**

-   Increasing Incidence of Diabetes world-wide.
    [@boultonGlobalBurdenDiabetic2005]

    -   2.8% in 2000, 4.4% in 2030

    -   25% of patients with diabetes develop a DFU at some stage in
        their lives

    -   Limb Loss every 20 seconds world-wide to Diabetes

**What's the pathophysiology of Diabetes and PAOD**
[@armstrong26PathophysiologyPrinciplesManagement2012]

-   Sensory Neuropathy, Motor Neuropathy and Autonomic Neuropathy

-   Structural and Gait abnormalities

-   Arterial disease

    -   Large Vessel

    -   Small Vessel

    -   Both

Given this diverse and confounding pathology the normal progressive
history of PAD is somewhat different. What's most concerning is the
neuropathy resulting in initial presentation being ulceration. This
results in a lack of a 'safety net' where presenting with progressive
claudication allows for a period of detection, management and and risk
factor modification before they develop tissue loss and are at risk of
amputation.

Since 2014 and the publication of WIfI a lot has changed in the way we
view PAD leading up to last years new Global Vascular Guidelines on
CLTI, which as a term has replaced CLI.
[@millsSocietyVascularSurgery2014a; @conteGlobalVascularGuidelines2019a]

## Intermittent Claudication

**What is Intermittent Claudication and the classic patient
presentation?**

The original population studies we mentioned determined the epidemiology
and natural history of Intermittent Claudication based on historically
validated and widely accepted questionnaires, namely the Rose
[@roseDiagnosisIschaemicHeart1962] (which later was adopted by the WHO)
and subsequently the Edinburgh questionnaire
[@lendEdinburghClaudicationQuestionnaire1992].

All questionnaires are based on a number of key diagnostic clinical
factors that define claudication, they are:

-   Onset

-   Calf involvement

-   Reproducibility

-   Relief with Rest

-   Not occurring at Rest

The progression historically graded by Fontaine (1954)
[@fontaineSurgicalTreatmentPeripheral1954] followed by the Rutherford
Grading System (1986, Revised 1997)
[@rutherfordRecommendedStandardsReports1997]

Rutherford et al. Ad Hoc Committee on Reporting Standards, SVS/North
American Chapter ISCVS:

+---------------+------------------------------------------------------+
| G             | Clinical Description                                 |
| rade/Category |                                                      |
+===============+======================================================+
| 0/0           | Asymptomatic -no haemodynamic significant occlusive  |
|               | disease                                              |
+---------------+------------------------------------------------------+
| I/1           | Mild Claudication                                    |
+---------------+------------------------------------------------------+
| I/2           | Moderate Claudication                                |
+---------------+------------------------------------------------------+
| I/3           | Severe Claudication                                  |
+---------------+------------------------------------------------------+
| II/4          | Ischaemic Rest Pain                                  |
+---------------+------------------------------------------------------+
| III/5         | Minor Tissue Loss                                    |
+---------------+------------------------------------------------------+
| III/6         | Major Tissue Loss                                    |
+---------------+------------------------------------------------------+

Differentiating neurogenic claudication from vasculogenic claudication
can be difficult. Vasculogenic claudication is releived by cessation of
ambulation and mostly affects a unilateral calf. Neurogenic improves
with postural changes, more often bilateral and affects the thighs.
[@nadeauReliabilityDifferentiatingNeurogenic2013]

**What is involved in the work up of patients with PAOD/Intermittent
Claudication?**

### History / Clinical Examination

SVS Guidelines:

"We recommend using ABI as the first-line non-invasive test to establish
a diagnosis of PAD in individuals with symptoms or signs suggestive of
disease. When the ABI is borderline or normal (\>0.9) and symptoms of
claudication are suggestive, we recommend an exercise ABI."

*Grade 1 Level of Evidence A*

-   ABPI

-   Exercise ABPI

-   Ultrasound

**What is an ABPI and how is it measured?**

The AHA came out with guidelines on how to perform an ABI and to
standardise the method to allow for more comparable results from
studies.

Divide the higher of the PT or DP pressure by the higher of the right or
left Brachial SBP *(Class 1 Level of Evidence A)*
[@aboyansMeasurementInterpretationAnkleBrachial2012]

Sensitivity and Specificity both \>95% (when ABPI cut off \</=0.9 -- in
detecting \>/= 50% stenosis) [@yaoAnkleSystolicPressure2005;
@ourielCriticalEvaluationStress1982]

| **Interpreting ABPI** |
|-----------------------|
| \>1.4                 |
| \>0.9-1.39            |
| 0.5-0.9               |
| 0.0-0.5               |

**What is Exercise ABPI studies?**

Constant Load Testing -- (unlike the Graded Test -- Bruce Protocol).

The mechanism for changes in ABPI are related to the fact that exercise
increases SBP and decreases SVR and therefore will affect flow across a
stenosis making it more physiologically relevant during exercise.
[@nicolaiReliabilityTreadmillTesting2009;
@alqahtaniPredictorsChangeAnkleBrachial2018]

Walking distance has been shown to correlate with level and severity of
POAD. [@strandnessHemodynamicsSurgeons1975a]

Patients with single level iliac disease may present with buttock/thigh
claudication with preserved pulses. Exercise treadmill ABI particularly
useful in determining the severity of disease in these patients.
[@aboyansGeneralPrognosisPatients2010]

**What is the ultrasound duplex criteria for defining PAOD?**

+-------------+--------------------+-------------+-------------------+
| Stenosis    | Peak Systolic      | Velocity    | Distal Artery     |
| Category    | Velocity           | Ratio       | Spectral Waveform |
+=============+====================+=============+===================+
| Normal      | \<150              | \<1.5       | Triphasic, Normal |
|             |                    |             | PSV               |
+-------------+--------------------+-------------+-------------------+
| 30-49%      | 150-200            | 1.5-2       | Triphasic, Normal |
|             |                    |             | PSV               |
+-------------+--------------------+-------------+-------------------+
| 50-75%      | 200-400            | 2-4         | M onophasic,      |
|             |                    |             | reduced PSV       |
+-------------+--------------------+-------------+-------------------+
| \>75%       | \>400              | \>4         | Damped, m         |
|             |                    |             | onophasic,        |
|             |                    |             | reduced PSV       |
+-------------+--------------------+-------------+-------------------+
| Occlusion   | No Flow -- B       |             |                   |
|             | -mode, Terminal    |             |                   |
|             | Thump              |             |                   |
+-------------+--------------------+-------------+-------------------+

Adapted from [@stone21VascularLaboratory2019]

**What Guidelines are there pertaining to PAD Management.**

SVS Guidelines (2015)

-   The SVS published the SVS practice guidelines for atherosclerotic
    occlusive disease of the lower extremities:**Management of
    asymptomatic disease and claudication**. *Conte and Pomposelli et
    al. JVS 2015* [@conteSocietyVascularSurgery2015]

Other Guidelines:

-   TASC 1 -2

-   European Guidelines (2017)

-   AHA Guidelines (last update 2016)

**What is the initial management of Asymptomatic Patients with PAD?**

1.  Smoking Cessation -- Multidisciplinary comprehensive smoking
    cessation interventions -- repeatedly until tobacco use has stopped
    (Grade A -- 1)
2.  Intervention is not only not recommended, but invasive treatment is
    recommended against, in the absence of symptoms (Grade A -1)

**How can we medically (non-invasively) manage Asymptomatic PAOD based
on the SVS Guidelines?**

-   Anti-platelet Therapy

    -   The Aspirin for Asymptomatic Atherosclerosis Trial -- n=3350,
        aspirin versus placebo. 8 years follow up no difference in
        events [@fowkesAspirinPreventionCardiovascular2010] -- therefore
        benefit unknown

-   Statin Therapy

    -   The Heart Protection Study
        [@heartprotectionstudycollaborativegroupRandomizedTrialEffects2007] -
        this study looked at Statins in patients with PAD but not
        completely asymptomatic, they had other risk factors such as
        diabetes, IHD, cerebral disease or hypertension. Without these
        risk factors Statin therapy benefit unsure.

    -   However, the AHA from the Framingham Study does recommend using
        Statins if 10-year risk based on risk calculators \>7.5% (which
        would be positive if PAD present).

-   Exercise and Limb Function

    -   No clear evidence that physical therapy improves QoL

-   Surveillance

    -   No benefit from US surveillance, unclear benefit of ABPI
        surveillance.

### Management

**How can we medically manage Intermittent Claudication based on the SVS
Guidelines?**

-   Smoking Cessation -- Multidisciplinary comprehensive smoking
    cessation interventions -- repeatedly until tobacco use has stopped
    (Grade A -- 1)

-   Dyslipidaemia: Statin Therapy Recommended -- most recent evidence on
    lipid therapy has suggested focusing on reducing 10-year
    cardiovascular event risk rather than specifically reducing lipid
    levels. (Grade 1-A)

-   Statin Therapy -- Aspirin therapy (75-325mg daily) is recommended to
    reduce cardiovascular events in patients with PAD (Grade 1-Level A)
    [@collaborationCollaborativeMetaanalysisRandomised2002]

    -   There is evidence that Clopidogrel 75mg compared to Aspirin is
        better in event reduction
        (CAPRIE)[@capriesteeringcommitteeRandomisedBlindedTrial1996]--
        replacing Aspirin with Clopidogrel Grade 1-Level B)

-   Diabetes Mellitus -- Optimization of HbA1C \< 7% (Grade 1-Level B)

-   Hypertension -- Indicated B-Blockers for hypertension (Grade 1-Level

    B)  (there's no evidence that Beta Blockers worsens IC)

-   Homocysteine -- Recommendation against Folic Acid and Vit B12 (
    Grade 2 -- C)

**To improve Limb Function in patients with IC:**

-   Cilostozol use -- IC without CHF -- 3-month Trial (Grade 2 - A)

    -   If unable to tolerate Cilostozol -- Pentoxifylline (400mg TDS)
        (Grade 2 -- B)

    -   Based on Meta-analysis 26 trials
        [@stevensSystematicReviewEfficacy2012]

-   Exercise Therapy

    -   First Line Therapy recommended SEP: minimum three times / week
        (30-60 min/session) for at least 12 weeks (Grade 1 Level A)

    -   Meta-analysis of 32 RCT's: Placebo versus exercise: Walking
        Time, Walking Ability, Pain Free Walking and maximum walking
        distance improves. BUT no difference in ABPI, Mortality or
        amputation. [@laneExerciseIntermittentClaudication2017]

    -   Meta- Analysis of 14RCT's: SEP better than Non-Supervised
        Programs. [@hagemanSupervisedExerciseTherapy2018]

**What is the management for patients with Intermittent Claudication?**

-   **Patient Selection for Intervention:**

    -   20-30% of patients with IC who adhere to risk factor
        modification will have progressive symptoms that will eventually
        be treated with intervention.

    -   Patient selection should be based on QoL and functional
        impairment in an active person (loss of ability to perform
        occupation or that limits basic activities of daily living)
        rather than haemodynamic (ABPI or US) or anatomical disease
        progression/severity.

-   **Always remember multifactorial causes of immobility --
    particularly in the elderly.**

    -   SVS recommends that invasive therapy for IC have a \>50%
        likelihood of sustained clinical improvement for at least 2
        years.

-   **Anatomical Selection:**

    -   Aortoiliac Disease:

        -   Previous TASC Classification has attempted to categorise
            anatomy of disease and subsequent recommendation of
            Endovascular versus open surgery. But as the authors of the
            SVS guidelines highlight, "improvements in technology and
            endovascular techniques have resulted in EVT replacing open
            surgical bypass as a primary treatment for both focal and
            advanced AIOD in many cases."

        -   The majority of evidence is non randomized and meta analyses
            of non-randomized series.

        -   Endovascular procedures over open surgery for focal AIOD
            causing IC. (Grade 1 Evidence B)

        -   Endovascular interventions as first line for CIA or EIZ
            occlusive disease-causing IC. (Grade 1 Level B)

        -   Hybrid recommended for Iliac disease involving CFA. (Grade 1
            Level B)

        -   Direct Surgical reconstruction (bypass, endarterectomy) in
            patients with reasonable surgical risk and diffuse AIOD not
            amenable to endovascular approach, after one or more failed
            attempts at EVT, or combined occlusive and aneurysmal
            disease. (Grade 1 Evidence B)

    -   Infrainguinal Disease:

        -   When you look at the historical data comparing all EVT
            together they are less durable than surgical bypass,
            especially when there's diffuse or long segments of
            occlusion/multilevel infra inguinal disease.

        -   Most recommendations are based on low level evidence when
            comparing EVT versus Open Surgery

        -   Focal + Not involving SFA origin = EVT (Grade 1 Level C)

        -   SFA 5-15cm, self-expanding stent (with or without
            paclitaxel) (Grade 1 Level B) -- NB: (This was in 2015 pre
            Katsanos Paper)

        -   Recommend against infrapopliteal treatment for IC (Grade 1
            Level C)

        -   Initial Surgical Bypass (with vein: Grade 1 Level a): If

            -   Diffuse FP disease

            -   Small Calibre \<5mm

            -   Extensive calcification in SFA

            -   Average or low operative risk (Grade 1 Level B)

## Chronic Limb Threatening Ischemia

**03 Jan 2021: *Nedal Katib and Danielle Bajakian***

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/PAOD_part2_postAuphonic.mp3")

```

**Since Our previous discussion of PAD how does this pathology then
progress clinically into the more advanced stages?**

The latter stages of both the Rutherford and Fontaine Classification
systems highlight this progression, with the Rutherford classification
of Stage 5 being specifically minor tissue loss with focal gangrene, and
stage 6 as major tissue loss identified by speeding of gangrene beyond
the Trans metatarsal level.

Rutherford et al. Ad Hoc Committee on Reporting Standards, SVS/North
American Chapter ISCVS:

+---------------+------------------------------------------------------+
| G             | Clinical Description                                 |
| rade/Category |                                                      |
+===============+======================================================+
| 0/0           | Asymptomatic -no haemodynamic significant occlusive  |
|               | disease                                              |
+---------------+------------------------------------------------------+
| I/1           | Mild Claudication                                    |
+---------------+------------------------------------------------------+
| I/2           | Moderate Claudication                                |
+---------------+------------------------------------------------------+
| I/3           | Severe Claudication                                  |
+---------------+------------------------------------------------------+
| II/4          | Ischaemic Rest Pain                                  |
+---------------+------------------------------------------------------+
| III/5         | Minor Tissue Loss                                    |
+---------------+------------------------------------------------------+
| III/6         | Major Tissue Loss                                    |
+---------------+------------------------------------------------------+

**What is Chrinc Limb Threatening Ischemia (CLTI), sometimes previously
call Critical Limb Ischemia (CLI)?**

In the last decade leading up to the recent Global Vascular Guidelines
(GVG) published last year, the term (Chronic Limb threatening Ischaemia)
CLTI has been gradually replacing CLI. The GVG mentions that their
"promotion" of the term CLTI is partly due to terms such as "critical or
severe limb ischemia" failing to "recognize the full spectrum and
inter-relatedness of components beyond ischemia that contribute to major
limb amputation..."

**What was the original definition and threshold for CLI and how can we
make sure we elicit the right symptoms from the patient?**

John Cranley back in his publication in 1969 defined Ischemic Rest Pain
as,

"...pain that occurs in the toes or in the area of the metatarsal heads.
Occasionally...in the foot proximal to the metatarsal heads. Elevation
of the limb above or at the horizontal position aggravates the pain and
pendency...brings relief..." [@cranleyIschemicRestPain1969]

Nocturnal Rest Pain: Worse due to horizontal positioning and systolic BP
drop during sleep.

**What aspects of the clinical assessment is important?**

Clinical Assessment involves a full history (the differential mentioned
above) and examination.

-   Clinical Examination:

    -   Bergers Test (Beurger 1908) / AKA Ratschows Test
        (Max-Ratschow-Klinic) identifies when there is critical ischemia
        without necrosis yet or gangrene, and is characterised by pallor
        when the leg is elevated above the level of the heart, which
        then turns red when hanging down over the edge of the bed. This
        redness is referred to as "Sunset appearance" and its due to
        abnormal autoregulation. Its been described that normally only a
        third of the capillary bed is open at any time but in a state of
        critical ischaemia because of the autoregulation being paralyzed
        a significantly higher portion of the capillaries open up.

    -   The ischaemic Angle: A refinement to Berger's Test: The angle of
        elevation from the horizontal at which the Doppler Signal of the
        PT or DP disappears. This is also referred to as the 'pole
        test', whereby the foot is raised alongside a calibrated pole
        marked in mmHg.

-   Tissue Loss:

    -   Gangrene Dry or Wet (infection)

    -   Level of tissue Loss

    -   Probing To Bone/ Exposed structures: Tendons, Soft Tissue, bone,
        Joint Capsule.

    -   Examination of an Ulcer (may have many aetiologies) -- important
        not only to identify extent of disease but also to exclude other
        aetiologies:

        -   Such as Venous, Mixed, infective, autoimmune, inflammatory,
            malignancy or trauma.

**What is the natural history of CLI and what do we know about its
prognosis?**

Fortunately, only a small portion of patients with Intermittent
Claudication will go on to develop rest pain or tissue loss. Its
estimated that anywhere between 5% -29% of patients with PAD or IC go on
to develop CLTI over 5 years.

However, those that do develop CLTI, have a high risk of limb loss
(greater than 20% annual risk) and a high mortality (10-15% annual
risk), the majority of terminal events being related to cardiovascular
events.

**What has changed in the last few decades?**

Prevalence of smokers-Ex and Current (decrease) and patients with
diabetes(increase).

Estimated 4.4% of the world will have diabetes by 2030. 25% of patients
with Diabetes will develop a foot ulcer.

**Briefly outline the underlying pathophysiology associated with
peripheral neuropathy in Diabetes?**

The loss of the basic nociceptive mechanisms in the foot amongst
diabetics, presents as a loss of protective sensation (LOPS).

Neuropathy can be divided into three types:

1.  Sensory: "stocking-glove" distribution

2.  Motor: Intrinsic muscle wasting -- resulting in deformities

3.  Autonomic: Sympathetic nervous system pathology

Along with Neuropathy, Diabetic Patients also have Structural
deformities and Gait disturbances in addition to Angiopathy or small
vessel disease.

There is often an overlap in the pathological processes in patients with
CLTI or CLI.

**What is the WIFI Classification?** [@millsSocietyVascularSurgery2014a]

Interestingly in the original article by Bob Rutherford regarding
Diabetes and PAD:

\- "It was generally agreed that diabetic patients who have a varied
clinical picture of neuropathy, ischemia and sepsis make the definition
even more difficult and it is desirable that these patients be
excluded...diabetic patients should be clearly defined as a separate
category or should be clearly defined as a separate category."

Since then, the SVS, while acknowledging that we can no longer exclude
these patients and treat them separately given the overlap, have decided
that a new classification system is necessary, as one of the key authors
(Joseph Mills) states:

-   "We classify things into groups to differentiate, remember and
    compare, observe and predict their behaviour over time." --Joseph
    Mills

WIfI stands for: Wound, Ischemia and foot Infection. Most of the
existing Vascular and non-Vascular classification systems don't include
all three components or fail to stratify the degree of ischemia and
presence of gangrene.

Principles of WIfI:

1.  Grades, Classes and Stages -- Each of the three categories (WIfI)
    have Grades 0,1,2,3: Resulting in 64 Classes.

2.  Delphi Consensus -- Clinical Stages 1 (Very Low),2 (Low), 3
    (Moderate), 4 (High Risk/Benefit).

    1.  What is the one-year risk of amputation with medical therapy
        alone?

    2.  What is the potential benefit from successful revascularization?

3.  Analogous to TNM Staging

"It is intended to be an iterative process with the goal of more
precisely stratifying patients according to their initial disease
burden, analogous to TNM cancer staging, but not to dictate therapy."

**What about a differential diagnosis or other causes of similar pain as
rest pain?**

Acute Lower Limb ischemia is a different clinical picture, but there may
some overlap with Acute on Chronic disease such as in the case of in
situ thrombosis in the lower limb arterial system.

Other causes of ischemic pain include:

-   Buergers Disease, or Thromboangiitis obliterans

-   Scleroderma

-   Fibromuscular dysplasia

-   Popliteal Artery Entrapment

-   Cystic Adventitial Disease

-   Persistent Sciatic Artery Disease

**What is the Rutherford Acute Ischemia Grading System?**

Although Acute Ischemia is very different from chronic ischemia,
patients with progressive chronic PAD can develop an acute picture
whether from embolism or in-situ thrombosis secondary to plaque rupture.
See \@ref(acute-limb-ischemia) for more.

**What are the recent CLTI Guidelines?**

In 2019 the SVS, the ESVS and the World federation of Vascular Societies
(WFVS) joined forces to put together the structure and funding of the
Global Vascular Guidelines Initiative (GVG). Importantly all sponsorship
was directly from the societies and any direct industry sponsorship or
external sources were excluded. They put together a steering committee
responsible for recruiting a large and diverse writing group and
outlined the scope and developed the section briefs of the guideline.

They determined that:

"The term"critical limb ischemia" (CLI) is outdated and fails to
encompass the full spectrum of patients who are evaluated and treated
for limb-threatening ischemia in modern practice."

CLTI was promoted as the term of choice and was defined by the target
population

The target population were:

1.  Ischemic Rest Pain with confirmatory hemodynamic studies

2.  Diabetic Foot Ulcer or any lower limb ulceration present for at
    least 2 weeks

3.  Gangrene involving any portion of the lower limb or foot

Exclusion from the population:

1.  Purely venous ulcers

2.  Acute Limb Ischemia/acute trash foot/ischemia due to emboli

3.  Acute Trauma or mangled extremity

4.  Wounds secondary to non-atherosclerotic conditions

Methodology of the guidelines utilised the structure of GRADE.

They Highlighted particular important sections in the evaluation and
management of patients with CLTI: namely Patient Risk stratification,
Limb Assessment and Severity of Limb Threat and the development of a
specific evidence-based revascularisation guideline in CLTI.

One think to notice (which the authors also highlight in the text body)
is compared to most guidelines, unfortunately in this area, particularly
when it comes to revascularisation, the level of evidence is generally
LOW. Again, highlighting the importance of these guidelines in
developing a standard approach and appropriately stratifying patients in
not only management but ongoing research.

**What clinical evaluation is necessary for the patient with CLTI?**

In addition to the History and Examination, and the WIfI assessment
mentioned above. For patients with diabetes and an ulcer a full
assessment of neuropathy and a "probe to bone" test for any open ulcers
is recommended as part of good practice.

We mentioned in our previous discussion the non-invasive methods of
assessment for these patients. One additional point to make is the
importance of TP and TBI in these patients.

It's been shown that, healing of an ulcer or tissue loss is unlikely if
a patient's toe pressures are less than 55mmHg. And Toe Pressures have
been validated in multiple studies to correlate with Amputation free
survival and wound healing: Amputation Free Survival TP \<30mmHg 2.13 HR
(1.52-2.98). - (J.E. Wickstrom et al EJVS May 2017).

**What are the recommendations for patients with CLTI when it comes to
Medical Therapy and Risk Factor Modification?**

-   Treat all patients with CLTI with an Antiplatelet agent (Grade 1
    Level A)

-   Consider Clopidogrel as the single agent (**Grade 2 Level B**) --
    CAPRIE

-   Moderate or High-intensity statin therapy to reduce all-cause and
    cardiovascular mortality (**Grade 1 Level A**)

-   Control Hypertension to BP target \<140mm Hg systolic and \<90mm Hg
    diastolic in patients with CLTI (**Grade 1 Level B**)

-   Offer Smoking Cessation interventions and ask all smokers or former
    smokers about status of tobacco use every visit (**Grade 1 Level
    A**)

**What imaging assessment is required?**

The CLTI Guidelines outlines an algorithm of attaining Arterial Anatomy
Imaging. Starting with US and then depending on the information
required, CTA, MRA or eventually digital subtraction angiography. They
emphasize the importance of obtaining good quality imaging to
appropriately stage and be able to compare the level and degree of
disease.

**What is the Global Limb Anatomic Staging System (GLASS)?**
[@conteGlobalVascularGuidelines2019a]

Because the existing arterial anatomical staging of disease is very
vague, and are "lesion focussed" and not all encompassing (beyond the
concept of 'in-line pulsatile flow to the foot'), GLASS attempts to
incorporate all aspects in its staging to improve vascular care and
evidence-based revascularisation (EBR) outcomes. 

GLASS incorporates two novel and important concepts: The Target Arterial
Path (TAP) and the estimated Limb-Based Patency (LBP) and it's a grading
system based on anatomical and subjective assessment of calcification.

GLASS focusses on Infrainguinal disease, with the Aorto Iliac (AI)
segment considered the inflow disease which includes the Common Femoral
Artery and the Profunda Artery. Therefore, the GLASS grades assume the
inflow vessels are treated and adequately 'dealt with'. 

Infrainguinal disease assessment for Femoropoliteal (FP) and
Infrapopliteal (IP) is based on length of disease and the extent of
CTO's. The FP and IP GLASS Grades are then combined into Stages 1-3. 

The calcification scale is a dichotomous subjective assessment of the
degree of calcification and if there is \>50% circumference of
calcification, diffuse or bulky calcification or "coral reef" plaques,
then there is an increase in the within-segment grade by one numerical
value.

There is also mention of the Inframaleolar (IM) degree of disease (PO,
P1-absent arch, P2-no target artery crossing into foot) which is not
included in the GLASS staging given little evidence on the outcomes this
difference makes on overall patency and limb salvage.

Once the GRADES (0-4) of FP and IP disease are determined then staging
(1-3) can be performed based on the matrix or grid that is provided.
Staging then allows for estimated Peripheral endo-Vascular Intervention
outcomes (PVI) to be predicted, Immediate Technical Failure (ITF - \<10%
or \< 20% or \> 20%)) and 1-year Limb Based Patency (LBP - \>70%, 50-75%
or \<50%). 

**What is the Target Arterial Path (TAP)?**

"The selected continuous route of in-line flow from groin to ankle. The
TAP typically involves the least diseased IP artery but may be angiosome
based."

**What revascularisation management strategies exist for CLTI?**

The mainstay of management for patients with CLI or CLTI has always been
based on the fundamental principle of limb salvage and given the high
risk of limb loss in these patients there's been a low threshold to
revascularize these patients if they have occlusive disease that is
treatable. But strategy has varied significantly.

The CLTI Guidelines provide an approach to dealing with this complex
condition on planning three aspects to each case:

1.  Patient Risk Estimation

2.  Limb Staging

3.  Anatomic Pattern of Disease

**What is involved with the Patient Risk Estimation?**

Good Practice Statements (Recommendations section 6)

"Refer all patients with suspected CLTI to a vascular specialist for
consideration of limb salvage, unless major amputation is considered
medically urgent."

"Offer primary amputation or palliation to patients with limited life
expectancy, poor functional status (e.g. non ambulatory), or an
unsalvageable limb after shared decision-making."

Recommendation 6.3:

-   Estimate periprocedural risk and life expectancy in patients with
    CLTI who are candidates for revascularization. **Grade 1 (Strong)
    Level of Evidence C (Low)**

-   Average Surgical Risk: \\\<5% operative mortality and 2-year
    survival \>50%

-   Severe Surgical Risk: \\\>/= 5% operative mortality and 2-year
    survival \</=50%

**What is involved and recommended with the Limb Staging and
recommendation for Management?**

Use an integrated threatened limb classification system (such as WIfI)
to stage all CLTI patients who are candidates for limb salvage. **Grade
1 (Strong) Level of Evidence C (Low)**

Perform urgent surgical drainage and debridement (including minor
amputation if needed) and commence antibiotic treatment in all patients
with suspected CLTI who present with deep space foot infection or wet
gangrene. (Good Practice Statement)

Offer Revascularisation to all "average surgical risk patients" (\\\<5%
operative mortality and 2-year survival \>50%)  with advanced
limb-threatening conditions (e.g. WIfI stage 4) and significant
perfusion deficits (e.g. ischemia grades 2 and 3). **Grade 1 (Strong)
Level of Evidence C (Low)**

**What is involved in the Planning of the Anatomic pattern of disease
and its effects of revascularisation strategy?**

The overall pattern of arterial occlusive disease is a dominant factor
in guiding type of revascularisation and timing of such.

**Do all patients require direct in-line flow to the foot as a primary
technical outcome with revascularisation?**

One important patient population to identify that do not necessarily
require direct in line flow are those with rest pain "for which
correction of inflow disease alone or treatment of FP disease even
without continuous tibial runoff to the foot may provide relief of
symptoms. This may also be the case in patients presenting with minor
degrees of tissue loss."

**What are some essential Key Factors to consider before deciding Open
versus Endovascular according to the CLTI guidelines?**

1.  **The "availability of and quality of autogenous vein conduit"7**

2.  **Patient overall risk (as mentioned above) and Limb Staging**

3.  **The Target EndoVascular Intervention (TVI) outcomes**

**What evidence do we have for deciding between Endo and Open? What is
the BASIL trial?**

The evidence is largely retrospective or non-controlled, industry
sponsored and in overall quality poor. BASIL (Bypass versus Angioplasty
in Severe Ischemia of the Leg) remains the only multicentre RCT (BASIL
-2 and 3 underway) directly comparing an endo versus open strategy in
CLTI and Infra-inguinal occlusive disease. 

BASIL compared POBA and Bypass across multiple centres (27 centres,
n=452, 1999-2004) in the UK. Primary endpoint was amputation-free
survival. [@bradburyBypassAngioplastySevere2005]

Major Findings:

1.  At 6-months follow up: no difference in AFS.

2.  Intention-To-Treat Analysis of overall follow up showed no
    significant difference in AFS and overall survival.

3.  Amongst patients who survived \>2 years, overall survival was better
    for those treated with Bypass as a first approach

4.  Analysis to treat: 

    1.  Prosthetic Bypass Patients did very poorly (even compared to
        POBA)  

    2.  Patients who had bypass after failed POBA had significantly
        worse AFS compared to those treated with a bypass as initial
        treatment

Criticism:

1.  Majority had POBA alone (not currently best endovascular option)

2.  25 % of Open Bypass were Prosthetic

3.  The Technology and Technical Skill with growing operator experience
    in Endovascular has improved.

**What is new with the future management guidance for CTLI? What is
BASIL 2 and 3 and BEST-CLI?**

BASIL 2: Infrapopliteal Disease: Vein Bypass First vs. Best Endovascular
Treatment first

BASIL 3: PBA +/- BMS vs. DCB +/- BMS vs. DES

(Both Follow up 24-60 months, Primary Endpoint AFS)

BEST-CLI:  Open Bypass versus Endovascular Intervention, Primary
Endpoint: MALE-Free Survival.  Major Above-the-Ankle Amputation, Major
Bypass or Jump/interposition graft revision or the need for thrombectomy
or thrombolysis (MALE).

## Acute Limb Ischemia

**17 Oct 2020:** *Alex Forsyth and Sarah Carlson; Boston Univeristy*

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/VSITE_ALI_postauphonic.mp3")

```

**What is acute limb ischemia and what does it encompass?**

Acute limb ischemia is defined as any process that leads to an abrupt
cessation of blood flow to a limb resulting in ischemia. There are a few
causes, but the most common two are embolic and thrombotic.

-   Embolic

    -   Cardiac

        -   Typically due to a fib

        -   Arm ischemia is most commonly due to cardiac embolism 

        -   Endocarditis -- as seen in IV drug users or patients with
            bacteremia from other causes

        -   Cardiac tumors -- such as atrial myxoma

    -   Atherosclerotic (e.g. iliac disease embolizing downstream to the
        lower leg)

    -   Paradoxical embolism (Thromboembolic venous system with PFO) 

    -   Aneurysm (e.g. thrombus from within an aortic aneurysm
        embolizing downstream to the leg)

-   Thrombosis

    -   Aneurysm

    -   Bypass graft

    -   Acute on chronic progression of atherosclerosis

-   Dissection

-   Thoracic outlet syndrome (in the upper extremity)

-   Vasospasm (severe)

For clarification: (1) Acute on chronic progression of atherosclerosis: 
-- e.g. once a chronic stenosis becomes critically tight, platelet
thrombus can develop leading to an acute occlusion; or unstable plaque
can "rupture" leading to an acute occlusion of a chronic lesion.  And
(2) Regarding aneurysms -- especially small aneurysms (such as
popliteal) -- these are less likely to rupture, but more likely to
thrombose and cause an acute limb ischemic event.

### Presentation and Diagnosis

**What is the patient presentation of ALI? Are there any differences for
upper vs. lower extremity presentations?**

Classically remembered by the 5 or 6 Ps depending on who you ask

-   Pain: usually located distal to the occlusion and gradually
    increases in severity as the duration of ischemia continues. The
    pain may also decrease after a time due to ischemic sensory loss

-   Pallor: the limb appears pale compared to the non ischemic limb.
    There is delayed capillary refill as well

-   Poikiolothermia: (just a way to make "cold" into a "P" -- really
    means cold limb) means literally the inability to regulate one's
    body temperature, or dependent on ambient temperature as cold
    blooded animals are. If there is no perfusion of warm blood to the
    limb, it acclimates to the ambient temperature. 

-   Pulseless: self explanatory, but a good thing to think about is if
    the contralateral limb has normal pulses, it suggests the absence of
    chronic limb ischemia and that an embolus or other cause of ALI. 

-   Paresthesia and Paralysis are the last two Ps. Paresthesias are an
    earlier sign of ischemic nerve dysfunction and paralysis is a later
    sign. In the lower extremity, ischemic changes often affect the
    anterior compartment first, and sensory loss over the dorsum of the
    foot is one of the earlier neurologic deficits in ALI

This is why a thorough physical exam is key; comparison of both limbs
and a good pulse exam including handheld doppler exam.  It can be
difficult for a junior resident to tell whether a limb is acutely
threatened, especially in patients with chronic disease where the
presentation of an acute change can be more subtle.  This is why the
attending surgeon will always ask the consult resident "how is the motor
and sensory function" in addition to the pulse exam...this helps us
gauge the chronicity and therefore the urgency of intervention.

**How is ALI classified?**

From Rutherford RB, Baker JD, Ernst C, et  al. Recommended standards for
reports dealing with lower extremity ischemia: revised version. J Vasc
Surg. 1997;26:517--538. [@rutherfordRecommendedStandardsReports1997]

+---------+---------+---------+---------+---------+---------+
| -   \   | **Pro g | S       | -   \*  | Arte    | Ve nous |
|     \*  | n       | ensory  |         | rial Si | Si gnal |
|     C   | osis**  | Loss    | Muscle  | gnal    |         |
|         |         |         | We a    |         |         |
| lassif  |         |         |         |         |         |
| i       |         |         | kne     |         |         |
|         |         |         | ss\*\*  |         |         |
| ca t i  |         |         |         |         |         |
| on\*\*  |         |         |         |         |         |
+=========+=========+=========+=========+=========+=========+
| I.      | Not imm | None    | None    | ++      | ++      |
|  Viable | e d     |         |         |         |         |
|         | iately  |         |         |         |         |
|         | th r e  |         |         |         |         |
|         | atened  |         |         |         |         |
+---------+---------+---------+---------+---------+---------+
| II\. T  | \       | \       | \       |         |         |
| h r e   |         |         |         |         |         |
| atened  |         |         |         |         |         |
+---------+---------+---------+---------+---------+---------+
| a\. M a | Sal v a | M       | None    | +/-     | \+      |
| r g     | geable  | inimal  |         |         |         |
| inally  | if p r  | (toes)  |         |         |         |
|         | omptly  | or none |         |         |         |
|         | t       |         |         |         |         |
|         | reated  |         |         |         |         |
+---------+---------+---------+---------+---------+---------+
| b\. Im  | Sal v a | More    | Mild, m | +/-     | \+      |
| m e d   | geable  | than    | o       |         |         |
| iately  | with i  | toes,   | derate  |         |         |
|         | m m     | as s o  |         |         |         |
|         | ediate  | ciated  |         |         |         |
|         | rev     | with    |         |         |         |
|         | ascula  | rest    |         |         |         |
|         | r i     | pain    |         |         |         |
|         | zation  |         |         |         |         |
+---------+---------+---------+---------+---------+---------+
| III\.   | Major   | P r o   | P r o   | \-      | \-      |
| Irr e v | tissue  | found,  | found,  |         |         |
| e       | loss or | an e s  | p a r   |         | |   v   |
| rsible  | p e r   | thetic  | alysis  |       | |         |
|         | manent  |         | ()      |         | |       |
| |       | nerve   |         | rigor)  | |       |         |
|         | d       |       | | \|      |         |         |
|     |   | amage\| |       | |         | C \|    |         |
|         | ine v i |         | |       |         |         |
|         | itable  | |       |         |         |         |
|         | \|C y   |         |         |         |         |
|         |         |         |         |         |         |
|         |         |       | |         |         |         |
+---------+---------+---------+---------+---------+---------+

         |yanosis|        |         |       |         |

|        \| may be\| \| \| \| \|
|        \| p \| \| \| \| \|
|         \|present \| \| \| \| \|

+--------+-------+---------+-------+---------+--------+

**What does the work up for an acute limb entail? How is the diagnosis
made?**

-   The diagnosis can often be made on history, physical exam, and
    bilateral ABIs. Imaging can be done in patients in who the diagnosis
    is uncertain

-   As with most urgent cases, the type of imaging done depends on the
    availability of at your institution, but generally imaging, such as
    a CT angiography or arteriography should be done on viable and
    marginally threatened limbs. Arteriography often can distinguish
    between embolic vs arterial thrombosis which may help to direct
    therapy.

The situation varies depending on how severe the presentation is and how
quickly you can obtain imaging.  Also depends on renal function and
whether you want accept two contrast loads (CT followed by endovascular
intervention).  As a rule of thumb, if I can feel femoral pulses I would
typically be more inclined to proceed with on-table angiogram without a
CT scan.  If femoral pulses are absent I would be more concerned about
aortoiliac disease and I would prefer to have a CT scan so I know what
I'm getting into in the operating room and can have a better plan.

In patients with severe renal insufficiency, MRA or MR time-of-flight
can be helpful, but these studies usually take a little longer to obtain
and may not be quickly available in an acute threatened limb situation.
Consider going straight to angiography, as to get the most immediate and
best imaging of the tibial vessels. [@creagerClinicalPracticeAcute2012;
@earnshaw100AcuteLimb2019]

Bedside ultrasound can also be very helpful even if you're not a
certified ultrasonographer yourself, if you have access to color flow
doppler US it can be very helpful.

### Management

**What is normally done in the initial management of ALI?**

-   Anticoagulation - IV unfractionated heparin is immediately
    administered to prevent proximal and distal progression of secondary
    thrombus as long as heparin is not contraindicated. The dose should
    be titrated to maintain activated partial thromboplastin time
    between 50 and 80 seconds (2-3 times normal values)

-   Supportive care (IV fluids)

-   A full set of labs including serum chemistry panel with BUN and Cr,
    CBC, baseline coagulation studies should be obtained. Baseline
    plasma CPK can be helpful to follow for evidence of rhabdomyolysis
    after reperfusion

A good rule of thumb for IV heparin is to start with a bolus of 80-100
units/kg, and then drop at 18units/kg/hr -- titrating to PTT at 2-3x
normal

**What are some of the options for treatment of ALI?**

-   Medical primarily with anticoagulation using heparin or a direct Xa
    inhibitor

-   Acute on chronic limb ischemia may be more technically challenging
    and require more complex reconstruction with a combination of
    endovascular and open techniques. [@dedonatoAcuteChronicLimb2018;
    @creagerClinicalPracticeAcute2012]

-   Open

    -   Thrombectomy -- balloon catheter based (Fogarty embolectomy
        balloon -- Dr. Fogarty invented this while he was a medical
        student)

        -   Patients with acute limb ischemia, neuro deficit and distal
            pop embolism may be best treated with popliteal exposure and
            open thrombectomy. [@darwoodSurgeryThrombolysisInitial2018;
            @kempeResultsSurgicalManagement2014]

    -   Bypass

        -   Patients with aortic occlusion and significant
            comorbidities, and significant aortoiliac baseline disease
            may likely benefit from primary axillo-bifemoral bypass.
            [@mohapatraRiskFactorsPerioperative2018]

    -   Endarterectomy -- not usually the go to but might be used for
        the common femoral 

-   Endovascular 

    -   Pharmacologic catheter directed thrombolysis

        -   Absolute contraindications include active bleeding disorder,
            CVA (\<6mo), CNS injury or head injury (\<3mo), or GI bleed
            (\<10d).

        -   Relative contraindications include recent major surgery,
            uncontrolled hypertension, intracranial tumor, pregnancy,
            recent eye surgery, hepatic failure, CPR (\<10d), or
            bacterial endocarditis.

        -   1-2% risk of hemorrhagic stroke. If neurologic deficit
            develops during thrombolysis, stop lysis and perform
            immediate head CT.
            [@ourielComparisonRecombinantUrokinase1998]

    -   Percutaneous thrombus aspiration -- useful for small fresh
        thrombi such as after angioplasty, as distal diameter of the
        catheter tip limits the size of the thrombus that can be
        removed. 

    -   Mechanical thrombolysis and aspiration are also useful for
        patients with contraindications for thrombolytic therapy, and
        also may allow for a lower dose of a thrombolytic agent, but
        risk damage to the arterial wall

**Who gets which kinds of treatment? Who needs emergent treatment?**

-   Class I might just need medical therapy like anticoagulation and
    revascularization can be elective. 

-   Class IIb patients do not need immediate revascularization

    -   If symptoms have been present for less than 2 weeks endovascular
        therapy is preferred

    -   If more than 2 weeks or lytic therapy has failed then surgical
        intervention is preferred 

-   Class IIb need immediate revascularization. Historically surgical
    revascularization has been preferred because of its immediacy, but
    catheter directed thrombolysis and percutaneous mechanical
    thrombectomy have shortened time to revascularization. 

-   Studies:

    -   STILE trial -- one of the first large RCTs comparing catheter
        thrombolysis with open surgery; overall the study showed some
        short term benefit to open surgery however this can probably be
        attributed to a couple things: (1) in 28% of patients randomized
        to CDT they weren't able to get a catheter in place so these
        patients were considered treatment failures and crossed over to
        the surgery arm, and (2) patients with very long durations of
        ischemia -- up to 6 months) were randomized, and when they
        looked at patients who had been symptomatic for less than 2
        weeks, the thrombolysis patients actually did better.
        [@thestileinvestigatorsResultsProspectiveRandomized1994]

    -   TOPAS trial -- larger RCT which enrolled patients who had an
        acute arterial occlusion of less than 14 days; this showed no
        difference in mortality or amputation-free survival but higher
        major bleeding in the CDT group.
        [@ourielComparisonRecombinantUrokinase1998]

    -   Meta analysis released originally published in 2002 but updated
        in 2013 and 2018 demonstrated no difference in mortality or limb
        salvage between surgical and thrombolytic therapy, but
        endovascular demonstrated higher rates of complications
        including ongoing limb ischemia and bleeding within 30 days of
        treatment. Previously they had reported higher rates of stroke
        in the thrombolysis category but the most recent update is
        unable to support this finding
        [@darwoodSurgeryThrombolysisInitial2018]

        -   All this to say, it is very reasonable to think about a
            cathter-directed therapy especially if the presentation is
            acute, less than 2 weeks or so.  That said, there are
            certain anatomical locations that most surgeons would favor
            a simple open procedure -- e.g. embolism to the common
            femoral or brachial arteries -- these are typically pretty
            simple to treat with a cut down and balloon thrombectomy.  

        -   One thing it is important to consider when doing a
            therapeutic infusion is that you might place a tPA infusion
            catheter at time zero and then bring the patient back 24
            hours later; patient needs to be advised that they'll need
            to lie flat for a day or even two days.  Setting
            expectations with patients is important.

-   Class III is usually treated with primary amputation because
    revascularization is unlikely to restore function to the limb and
    restoring bloodflow can cause the patient serious harm. 

**What are the risks of revascularization for a class III or prolonged
ischemia?**

-   Myonephropathic metabolic syndrome: muscle cells undergo
    liquefactive necrosis due to ischemia. Potassium, myoglobin, lactic
    acid, and superoxide accumulate and can perfuse through the body or
    can have a sudden increase in the event of revascularization which
    leads to hyperkalemia, arrhythmias, pulmonary edema, metabolic
    acidosis, myoglobinuria, and can even cause sudden death from heart
    and/or renal failure. [@obaraAcuteLimbIschemia2018] Treatment of
    this is largely supportive with fluids

### Compartment Syndrome

**What is the pathophysiology and manifestations of compartment
syndrome?**

-   Increased intramuscular compartment pressure results from increases
    in capillary permeability due to ischemic reperfusion. The increase
    in pressure leads to neuromuscular dysfunction and interferes with
    circulation. Irreversible damage occurs when pressures exceed 30mmHg
    in each compartment  

-   Clinically they can have neurological dysfunction with sensory motor
    deficits, but the most common presentation is a tense extremity with
    pain on passive movement of the muscles in the compartment, which is
    often dorsiflexion/plantar flection of the ankle. A sensitive
    indicator is loss of two point discrimination

**How do you diagnose a compartment syndrome?**

-   Physical exam (tenderness, especially over anterior compartment),
    paresthesias, especially between first and second toes (anterior
    compartment:  deep peroneal nerve -- this is a VSITE favorite)

-   Compartment pressures:  how do you do this?

    -   Need a needle to access the compartment and a pressure
        monitoring system (can be handheld Stryker kit, or just a hollow
        bore needle connected to an arterial pressure bag).

    -   Normal compartment pressure is \<10-20mmHg; greater than 30 is
        highly concerning.  Probably even more accurate than an absolute
        number is comparing the compartment pressure to the mean
        arterial pressure or diastolic pressure.  If the compartment
        pressure is within 40mmHg of the MAP (for example, MAP is 60 and
        compartment pressure is 25 -- this is concerning) -- OR -- if
        the difference between compartment pressure and diastolic
        pressure is less than 10 (for example, diastolic pressure is low
        at 30, and compartment pressure is 22).

If clinical suspicion of compartment syndrome is high, I tend not to be
reassured by "normal" compartment pressures.  It's relatively low risk
to do fasciotomies, but the risk of limb loss is so high for a missed
compartment syndrome...I would much rather err on the side of caution if
there's any question.

**Treatment is a fasciotomy -- can you tell us a little about the types
of fasciotomies?**

-   Forearm and upper arm fasciotomies are often performed by orthopedic
    or hand surgeons. The forearm fasciotomy includes dorsal and volar
    incisions to release the dorsal and volar compartment, and mobile
    wad while avoiding numerous superficial cutaneous nerves. The arm
    fasciotomy releases the medial, lateral, and deltoid compartments
    through medial lateral incisions 

-   Fasciotomies can be done in the thigh as well with a medial and
    lateral incision to release the lateral, medial and posterior
    compartment

-   The most common type is a lower leg 4 compartment fasciotomy 

**How is a lower leg 4 compartment fasciotomy for the lower extremity
performed?**

1.  A longitudinal incision is created between the fibular shaft and the
    crest of the tibia over the intermuscular septum and the anterior
    and lateral compartments are opened. If tissues are swollen
    occluding the view of the intermuscular septum, the perforating
    vessels can be followed down to it. Nerves including the peroneal
    nerve are most at risk near the fibular head

2.  A second incision is created on the medial surface of the lower leg
    approx. 1cm posterior to the edge of the tibia to avoid the greater
    saphenous vein. The superficial posterior compartment is incised.
    The gastrocnemius-soleus complex is taken down from its attachments
    to the tibia in order to access the deep posterior compartment. 

3.  The incisions are made generously -- sometimes the skin incision can
    be a little short of the fascial incision, but they should be nice
    and long in order to fully release the compartments

4.  After hemostasis be sure to apply loose dressings, and the leg
    should be elevated to reduce edema that can complicate closure.
    Closure can be done in 48-72 hours but may be delayed and dressed
    with wound vacs to attempt primary closure. If primary closure is
    not possible, a split thickness skin graft can be used for closure 

**Who should a prophylactic fasciotomy be performed on?**

-   Patients with high occlusion and extensive ischemia, 

-   Acute ischemia of greater than 6 hours with few collaterals

-   Patients with combined arterial and venous injury

-   Patients who are obtunded making serial examination difficult

**What is the prognosis for a patient with ALI? What are some patient
factors that lead to a poor prognosis?**

Amputation rates after acute limb ischemia are typically described in
the 10-20% range, and mortality is also in the 10-25% range whether
you're talking about surgery or catheter-directed procedures (that's
excluding the patients who present with Rutherford class III and by
definition have an unsalvageable limb).  Many factors determine
likelihood of amputation; typically, patients with more medical comorbid
conditions tend to do worse as you might expect:  baseline CAD, kidney
disease and smoking are predictive of worse outcomes.  There is a trend
toward improved limb salvage rates (decreased amputation rate) over
time, and I think this speaks to wider availability of different limb
salvage techniques among vascular surgeons across the globe.

## Non-atherosclerotic occlusive disease

### Blue Toe Syndrome

#### Etiology and Presentation

Blue toe syndrome is characterized as a painful discolored toe in the
setting of a normal vascular examination or palpable pulses.

Acute onset of blue toe after MI - consider cholesterol embolization
from wire manipulation in the aorta.
[@saricAorticAtherosclerosisEmbolic2012]

#### Evaluation

Patients should have a work up for an embolic source. Patients without a
clear recent source of embolization on history or physical, should
undergo CTA of chest/abdomen/pelvis to look for an arterial lesion,
echocardiogram to identify a cardiac source, and a Holter monitor to
evaluate for an underlying arrhythmias.

If no obvious source of embolism can be identified, further evaluation
for underlying cancer may lead to a prothrombotic state.

#### Management

Treatment of blue toe from cholesterol embolization best with single
antiplatelet and statin. Anticoagulation may precipitate further
embolization. [@ghahramaniPostproceduralBlueToes2016;
@quinonesCholesterolEmboliSyndrome2013]

Treatment of blue toe from a specific thrombogenic arterial lesion may
be best with anticoagulation. Recurrent symptoms on anticoagulation may
require stent coverage or, in a young patient, may require thrombectomy.
[@reyesvaldiviaAnticoagulationAloneAortic2017;
@jeyabalanEndovascularStrategiesTreatment2014;
@vermaContemporaryManagementSymptomatic2014]\

### Popliteal entrapment syndrome

#### Etiology and Presentation

Most often seen in young athletes with embryologically abnormal lateral
attachment of medial head of the gastrocnemius muscle.[@gokkus2014;
@lejay2014]

#### Evaluation

Physical examination finding will be loss of pedal pulse on active
plantarflexion or passive dorsiflexion of the foot.[@gokkus2014;
@lejay2014]

#### Management 
