---
editor_options: 
  markdown: 
    wrap: 72
---

# Vascular Trauma

## Peripheral

**22 Sep 2019**: *Kevin Kniery, MD, MPH; Todd Rasmussen, MD*

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/Rasmussen_Final_Final.mp3")

```

## Abdominal Arterial

**22 Dec 2019:** *Kevin Kniery, MD, MPH; Adham Elmously, MD; Todd
Rasmussen, MD*

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/Abdominal_Trauma_Part_One.mp3")

```

## Abdominal Venous

**18 Jan 2020**: *Kevin Kniery, MD, MPH; Adham Elmously, MD; Todd
Rasmussen, MD*

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/Vascular_Venous_Auphonic.mp3")

```

Check out Dr. Rasmussen's book

<https://www.elsevier.com/books/richs-vascular-trauma/9781455712618>

<https://www.amazon.com/Richs-Vascular-Trauma-Todd-Rasmussen/dp/1455712612>

## Cerebrovascular Trauma

**25 May 2020**: *Kevin Kniery, MD, MPH; Adam Johnson, MD, MPH; Nicole
Rich, MD, MPH; Todd Rasmussen, MD*

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/Cerbrovascular_Trauma_auphonic.mp3")

```

EAST Blunt Cerebrovascular Injury Guidelines

<https://www.east.org/education/practice-management-guidelines/blunt-cerebrovascular-injury> 

Check out Rich's Vascular Trauma

<https://www.elsevier.com/books/richs-vascular-trauma/9781455712618>

## Endovascular Approaches

**09 Feb 2021**: *Kevin Kniery, MD, MPH; Marlin "Wayne" Causey MD; Todd
Rasmussen, MD*

```{r echo=FALSE}

embed_audio("https://traffic.libsyn.com/audiblebleeding/Rasmussen_Endovascular_Trauma.mp3")

```

Seminal Papers in Blunt Thoracic Aortic Injury

-   AAST 1997 Paper: <https://pubmed.ncbi.nlm.nih.gov/9095103/>

-   AAST 2008 Paper: <https://pubmed.ncbi.nlm.nih.gov/18545103/>

-   JVS 2011 Paper: <https://pubmed.ncbi.nlm.nih.gov/20974523/>

-   Timing of repair of BTAI JVS
    2020: <https://www.jvascsurg.org/article/S0741-5214(20)31575-5/fulltext>

Dr. Ben Starnes' podcast on Behind The Knife on
BTAI: <https://bit.ly/2LuycWq>
