--- 
title: "Vascular Surgery Exam Prep"
author: "Editors: Adam Johnson, MD, MPH; Matt Smith, MD, PhD; and Audible Bleeding"
date: "DRAFT - `r Sys.Date()`"
output: pdf_document
documentclass: book
bibliography: [references.bib, packages.bib]
csl: apa-numeric-superscript-brackets.csl
url: https://adam-mdmph.github.io/VS-Exam-Prep/
description: |
  This is an ebook version of exam preparation content produced for Audible Bleeding: The Vascular Surgery Podcast, powered by the Society for Vascular Surgery
link-citations: yes
github-repo: https://github.com/adam-mdmph/VS-Exam-Prep
site: bookdown::bookdown_site
apple-touch-icon: "images/apple_icon.png"
---


```{r message=FALSE, warning=FALSE, include=FALSE}
library(devtools)

devtools::install_github("mccarthy-m-g/embedr")
library(embedr)
```


# About

This content was developed by the [Audible Bleeding Team](https://www.audiblebleeding.com/about-1/) to accompany our board review podcast series.
Each chapter covers a key domain of vascular surgery, essential for board review preparation, and is associated with an Audible Bleeding episode which you can access through the link embeded into the text. 

## Usage 

This e-book is not intended to be a comprehensive board review guide, but instead serves as an easily accessible resource, paired with our podcast content to help streamline exam preparation.

## Comments, Questions or Contributions

Please visit our [github page](https://github.com/adam-mdmph/VS-Exam-Prep) or [send us an email](mailto:audiblebleeding@vascularsociety.org).


```{r include=FALSE}
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'embedr'
), 'packages.bib')
```
